---
title: "Two-Sample Test - HSC-Ery"
author: "Mervin M Fansler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    code_folding: show
    toc: true
    toc_float: true
---

# Purpose

This document demonstrates the use of a two-sample bootstrap test using the 
HSPC data from *Dahlin et al., 2018* and *Wolf et al., 2019*.

# Initialization

## Libraries
```{r libs, message=FALSE, warning=FALSE}
library(magrittr)
library(tidyverse)
library(cowplot)
library(SingleCellExperiment)
library(scutrboot)
```

## Parameters
```{r set_params}
set.seed(20230522)
N_BOOTSTRAPS=10000
MIN_NCELLS=50
```

## Functions

```{r methods}
plot_pval_histogram <- function (df) {
  ggplot(df, aes(x=pval)) +
    geom_histogram(boundary=0, bins=41, 
                   color='black', fill='lightgrey', linewidth=0.2) +
    scale_x_continuous(expand=c(0.02,0.02)) +
    scale_y_continuous(expand=c(0,0,0,100)) +
    labs(x="p-values", y="Genes") +
    theme_bw()
}
```

# Data
## Loading
```{r load_data, message=FALSE}
sce <- readRDS("data/sce/hspcs.txs.full_annot.Rds")
```

## Preprocessing
```{r prepare_data}
sce %<>% `[`(!rowData(.)$is_blacklisted,)
idx_multi <- rowData(sce)$atlas.utr_type == 'multi'
idx_consistent <- rowData(sce)$is_consistent

gene_id2name <- rowData(sce) %>% as_tibble %>% pull(gene_name, name=gene_id)
```

# Analysis

## WD Test: Ery vs HSC

```{r test_ery_hsc_wd}
df_test_ery_hsc_wd <- testTwoSample(sce[idx_multi,], assayName="normcounts",
                                    sampleKey="clusters", geneKey="gene_id",
                                    sample0="14/Stem", sample1="12/Ery", 
                                    statistic="WD", nBootstraps=N_BOOTSTRAPS, 
                                    minCellsPerGene=MIN_NCELLS, 
                                    featureExclude="is_ipa", 
                                    ) %>%
  as_tibble() %>%
  mutate(gene_name=gene_id2name[gene],
         qval=p.adjust(pval, method="BH")) %>%
  select(gene_name, everything())
```

### P-Values Plot
```{r plot_pvals_wd, fig.width=4, fig.height=2}
plot_pval_histogram(df_test_ery_hsc_wd)

ggsave("output/figure4/sup-pvals-ery-hsc-wd.pdf", width=4, height=2, dpi=300)
```

### Export
```{r export_wd}
sprintf("data/dtu/%s-hsc-ery-twosample-wd.min50.boot10k.tsv.gz", 
        format(Sys.time(), '%Y%m%d')) %>%
  write_tsv(x=df_test_ery_hsc_wd)
```

## LUI Test: Ery vs HSC

```{r test_ery_hsc_lui}
df_test_ery_hsc_lui <- testTwoSample(sce[idx_multi & idx_consistent,], assayName="normcounts",
                                    sampleKey="clusters", geneKey="gene_id",
                                    sample0="14/Stem", sample1="12/Ery", 
                                    statistic="UI", featureIndex="is_distal",
                                    nBootstraps=N_BOOTSTRAPS, 
                                    minCellsPerGene=MIN_NCELLS, 
                                    featureExclude="is_ipa") %>%
  as_tibble() %>%
  mutate(gene_name=gene_id2name[gene],
         qval=p.adjust(pval, method="BH")) %>%
  select(gene_name, everything())
```

### P-Values Plot
```{r plot_pvals_lui, fig.width=4, fig.height=2}
plot_pval_histogram(df_test_ery_hsc_lui)

ggsave("output/figure4/sup-pvals-ery-hsc-lui.pdf", width=4, height=2, dpi=300)
```

### Export
```{r export_lui}
sprintf("data/dtu/%s-hsc-ery-twosample-lui.min50.boot10k.tsv.gz", 
        format(Sys.time(), '%Y%m%d')) %>%
  write_tsv(x=df_test_ery_hsc_lui)
```

## WUI Test: Ery vs HSC

```{r test_ery_hsc_wui}
sce_subset <- sce[idx_multi,]

rowData(sce_subset) %<>% as_tibble() %>%
  group_by(gene_id) %>%
  mutate(utr_rank=rank(utr_position),
         utr_wt=(utr_rank - 1)/(max(utr_rank) - 1)) %>%
  ungroup() %>%
  as.data.frame() %>%
  DataFrame(row.names=.$transcript_id) %T>%
  { stopifnot(all(rownames(.) == rownames(sce_subset))) }

df_test_ery_hsc_wui <- testTwoSample(sce_subset, assayName="normcounts",
                                    sampleKey="clusters", geneKey="gene_id",
                                    sample0="14/Stem", sample1="12/Ery", 
                                    statistic="WUI", featureIndex="utr_wt",
                                    nBootstraps=N_BOOTSTRAPS, 
                                    minCellsPerGene=MIN_NCELLS, 
                                    featureExclude="is_ipa") %>%
  as_tibble() %>%
  mutate(gene_name=gene_id2name[gene],
         qval=p.adjust(pval, method="BH")) %>%
  select(gene_name, everything())
```

### P-Values Plot
```{r plot_pvals_wui, fig.width=4, fig.height=2}
plot_pval_histogram(df_test_ery_hsc_wui)

ggsave("output/figure4/sup-pvals-ery-hsc-wui.pdf", width=4, height=2, dpi=300)
```

### Export
```{r export_wui}
sprintf("data/dtu/%s-hsc-ery-twosample-wui.min50.boot10k.tsv.gz", 
        format(Sys.time(), '%Y%m%d')) %>%
  write_tsv(x=df_test_ery_hsc_wui)
```

## IPA Test: Ery vs HSC

```{r test_ery_hsc_ipa}
idx_ipa <- rowData(sce) %>%
  as_tibble %>%
  group_by(gene_id) %>%
  mutate(has_ipa=any(is_ipa), has_nipa=any(!is_ipa)) %>%
  ungroup() %>%
  select(transcript_id, is_ipa, has_ipa, has_nipa) %>%
  filter(has_ipa, has_nipa) %$%
  transcript_id

df_test_ery_hsc_ipa <- testTwoSample(sce[idx_ipa,], assayName="normcounts",
                                    sampleKey="clusters", geneKey="gene_id",
                                    sample0="14/Stem", sample1="12/Ery", 
                                    statistic="UI", featureIndex="is_ipa",
                                    nBootstraps=N_BOOTSTRAPS, 
                                    minCellsPerGene=MIN_NCELLS) %>%
  as_tibble() %>%
  mutate(gene_name=gene_id2name[gene],
         qval=p.adjust(pval, method="BH")) %>%
  select(gene_name, everything())
```

### P-Values Plot
```{r plot_pvals_ipa, fig.width=4, fig.height=2}
plot_pval_histogram(df_test_ery_hsc_ipa)

ggsave("output/figure4/sup-pvals-ery-hsc-ipa.pdf", width=4, height=2, dpi=300)
```

### Export
```{r export_ipa}
sprintf("data/dtu/%s-hsc-ery-twosample-ipa.min50.boot10k.tsv.gz", 
        format(Sys.time(), '%Y%m%d')) %>%
  write_tsv(x=df_test_ery_hsc_ipa)
```

---

# Runtime Details
## Session Info
```{r sesh_info, echo=FALSE}
sessionInfo()
```

## Conda Environment
```{bash comment="", echo=FALSE}
if ! command -v conda &> /dev/null
then
  echo "Conda not detected."
elif [ -z "${CONDA_PREFIX}" ]
then
  echo "No active Conda environment."
else
  echo "## Conda Environment YAML"
  conda env export
fi
```
