---
title: "Expanded Cell Types - DGE vs DTU Tests"
author: "Mervin M Fansler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    code_folding: show
    toc: true
    toc_float: true
---

# Purpose

This compares the overlap of significant genes detected by differential gene 
expression testing versus differential isoform usage testing.

# Initialization

## Libraries
```{r libs, message=FALSE, warning=FALSE}
library(magrittr)
library(tidyverse)
library(writexl)
library(cowplot)
```

## Parameters
```{r set_params}
MAX_QVAL=0.05
MIN_L2FC=log2(1.5)
MIN_WD=0.15
```

# Data
## Loading
```{r load_data, message=FALSE}
df_dtu <- read_tsv("data/dtu/20211026-diff-pairwise-wd-tests.min50.boot10k.tsv.gz") %>%
  dplyr::rename(gene_id=gene)
df_dge <- readRDS("data/dge/merged.pairwise_DEGs.rds") %>%
  rename(start=cell1, end=cell2)
```

## Preprocessing
```{r prepare_data}
df_sig_dtu <- df_dtu %>%
  mutate(sig_isoform=qval < MAX_QVAL & abs(stat) >= MIN_WD) %>%
  select(gene_id, start, end, sig_isoform)

df_sig_dge <- df_dge %>%
  mutate(sig_gene=p_val_adj < MAX_QVAL & abs(avg_log2FC) >= MIN_L2FC) %>%
  select(gene_id, start, end, sig_gene)

df_sig_all <- full_join(df_sig_dtu, df_sig_dge, by=c("gene_id", "start", "end")) %>%
  replace_na(list(sig_isoform=FALSE, sig_gene=FALSE)) %>%
  mutate(sig=ifelse(sig_isoform & sig_gene, "Both", 
                    ifelse(sig_isoform, "DTU", 
                           ifelse(sig_gene, "DGE", "NS"))),
         sig=factor(sig, levels=c("NS", "DTU", "Both", "DGE")),
         test_label=str_c(end, " vs ", start))

df_sig_multi <- left_join(df_sig_dtu, df_sig_dge, by=c("gene_id", "start", "end")) %>%
  replace_na(list(sig_isoform=FALSE, sig_gene=FALSE)) %>%
  mutate(sig=ifelse(sig_isoform & sig_gene, "Both", 
                    ifelse(sig_isoform, "DTU",
                           ifelse(sig_gene, "DGE", "NS"))),
         sig=factor(sig, levels=c("NS", "DTU", "Both", "DGE")),
         test_label=str_c(end, " vs ", start))
```

# All Genes
## Everything
```{r plot_all_ns, fig.width=6, fig.height=5}
colvals_ns <- c(NS="#CCCCCC", DTU="#EEE643", Both="#2E9F72", DGE="#2170B1") %>% {
  setNames(colorspace::desaturate(., 0.2), names(.))
}

df_sig_all %>%
  mutate(test_label=fct_reorder(test_label, sig, . %>% { sum(. != "NS") }, .desc=TRUE)) %>%
  ggplot(aes(x=test_label, fill=sig)) +
  geom_bar(position='stack', color='black', size=0.1) +
  scale_fill_manual(breaks=names(colvals_ns), values=colvals_ns) +
  scale_x_discrete(guide=guide_axis(angle=45)) +
  scale_y_continuous(expand=c(0,0,0.05,0)) +
  labs(x="Compared Cell Types", y="All Tested Genes", fill="Difference") +
  theme_bw()

#ggsave("output/figure4/fig4f-celltypes-dge-dtu-compared-all-ns.pdf", width=6, height=5, dpi=300)
```

## Plot Positive
```{r plot_all, fig.width=6, fig.height=5}
colvals <- c(DTU="#EEE643", Both="#2E9F72", DGE="#2170B1") %>% {
  setNames(colorspace::desaturate(., 0.2), names(.))
}

df_sig_all %>%
  filter(sig != "NS") %>%
  ggplot(aes(x=fct_infreq(test_label), fill=sig)) +
  geom_bar(position='stack', color='black', size=0.1) +
  scale_fill_manual(values=colvals) +
  scale_x_discrete(guide=guide_axis(angle=45)) +
  scale_y_continuous(expand=c(0,0,0.05,0)) +
  labs(x="Compared Cell Types", y="Significant Genes", fill="Difference") +
  theme_bw()

#ggsave("output/figure4/fig4f-celltypes-dge-dtu-compared-all.pdf", width=6, height=5, dpi=300)
```

## Table 
```{r tbl_all_genes}
df_sig_all %>%
  group_by(start, end) %>%
  dplyr::count(sig_gene, sig_isoform) %>%
  knitr::kable()
```

# Multi-UTR Genes
## Plot Positive-Only
```{r plot_multi, fig.width=6, fig.height=5}
df_sig_multi %>%
  filter(sig != "NS") %>%
  ggplot(aes(x=fct_infreq(test_label), fill=sig)) +
  geom_bar(position='stack', color='black', size=0.1) +
  scale_fill_manual(breaks=names(colvals), values=colvals) +
  scale_x_discrete(guide=guide_axis(angle=45)) +
  scale_y_continuous(expand=c(0,0,0.05,0)) +
  labs(x="Compared Cell Types", y="Significant Multi-UTR Genes", 
       fill="Difference") +
  theme_bw()

#ggsave("output/figure4/fig4f-celltypes-dge-dtu-compared-multi.pdf", width=6, height=5, dpi=300)
```

## Plot All
```{r plot_multi_ns, fig.width=6, fig.height=5}
colvals_ns <- c(NS="#CCCCCC", DTU="#EEE643", Both="#2E9F72", DGE="#2170B1") %>% {
  setNames(colorspace::desaturate(., 0.2), names(.))
}

df_sig_multi %>%
  mutate(test_label=fct_reorder(test_label, sig, . %>% { sum(.=="DTU") }, .desc=TRUE)) %>%
  ggplot(aes(x=test_label, fill=sig)) +
  geom_bar(position='stack', color='black', size=0.1) +
  scale_fill_manual(breaks=names(colvals_ns), values=colvals_ns) +
  scale_x_discrete(guide=guide_axis(angle=45)) +
  scale_y_continuous(expand=c(0,0,0.05,0)) +
  labs(x="Compared Cell Types", y="Coexpressed Multi-UTR Genes", 
       fill="Difference") +
  theme_bw()

#ggsave("output/figure4/fig4f-celltypes-dge-dtu-compared-multi-ns.pdf", width=6, height=5, dpi=300)
```

## Table 
```{r tbl_multi_genes}
df_sig_multi %>%
  group_by(start, end) %>%
  dplyr::count(sig_gene, sig_isoform) %>%
  knitr::kable()
```

## Chi-Square Tests
```{r}
df_sig_multi %>%
  group_by(start, end) %>%
  nest() %>%
  transmute(start, end, res=map(data, ~ broom::tidy(chisq.test(.x$sig_gene, .x$sig_isoform, simulate.p.value=TRUE)))) %>%
  unnest(res) %>%
  select(start, end, statistic, p.value) %>%
  knitr::kable()
```


```{r tbl_wide}
df_chisq <- df_sig_multi %>%
  group_by(start, end) %>%
  nest() %>%
  transmute(start, end, res=map(data, ~ broom::tidy(chisq.test(.x$sig_gene, .x$sig_isoform, simulate.p.value=TRUE, B=2000)))) %>%
  unnest(res) %>%
  select(start, end, statistic, p.value)

map_sig <- c("FALSE_FALSE"="NS",
             "FALSE_TRUE"="DTU",
             "TRUE_FALSE"="DGE",
             "TRUE_TRUE"="Both")

df_wide <- df_sig_multi %>%
  group_by(start, end) %>%
  dplyr::count(sig_gene, sig_isoform) %>%
  mutate(result=map_sig[str_c(sig_gene,sig_isoform, sep="_")]) %>%
  ungroup() %>%
  pivot_wider(id_cols=c("start", "end"), names_from="result", values_from="n", values_fill=0L) %>%
  mutate("Coexpressed multi-UTR genes"=NS+DTU+DGE+Both) %>%
  left_join(df_chisq, by=c("start", "end")) %>%
  rename("Cell Type 1"=start,
         "Cell Type 2"=end,
         "Chi-squared"=statistic,
         "p-value"=p.value) %>%
  select(`Cell Type 1`, `Cell Type 2`, `Coexpressed multi-UTR genes`,
         NS:Both, `Chi-squared`, `p-value`)

knitr::kable(df_wide)
```

## Export Table
```{r export_tbl}
write_xlsx(df_wide, "output/tables/tbl3-extended-celltypes-chisq-results-mast.xlsx")
```

---

# Runtime Details
## Session Info
```{r sesh_info, echo=FALSE}
sessionInfo()
```

## Conda Environment
```{bash comment="", echo=FALSE}
if ! command -v conda &> /dev/null
then
  echo "Conda not detected."
elif [ -z "${CONDA_PREFIX}" ]
then
  echo "No active Conda environment."
else
  echo "## Conda Environment YAML"
  conda env export
fi
```
