---
title: "Ery Trajectory - DGE vs DTU Tests"
author: "Mervin M Fansler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    code_folding: show
    toc: true
    toc_float: true
---

# Purpose

This compares the overlap of significant genes detected by differential gene 
expression testing versus differential isoform usage testing.

# Initialization

## Libraries
```{r libs, message=FALSE, warning=FALSE}
library(magrittr)
library(tidyverse)
library(cowplot)
```

## Parameters
```{r set_params}
MAX_QVAL=0.05
MIN_L2FC=log2(1.5)
MIN_DLUI=0.15
```

# Data
## Loading
```{r load_data, message=FALSE}
df_dtu <- read_tsv("data/dtu/20211019-ery-trajectory-pairwise-lui-tests.min50.boot10k.tsv.gz",
                   col_types='ccccddi') %>%
  dplyr::rename(gene_id=gene)
df_dge <- read_tsv("data/dge/20211026-ery-trajectory-pairwise-t-tests.tsv.gz",
                   col_types='ccccddd')
```

## Preprocessing
```{r prepare_data}
df_sig_dtu <- df_dtu %>%
  mutate(qval=p.adjust(pval, method="BH"),
         sig_isoform=qval < MAX_QVAL & abs(stat) >= MIN_DLUI) %>%
  group_by(gene_id) %>%
  summarize(sig_isoform=any(sig_isoform))

df_sig_dge <- df_dge %>%
  mutate(qval=p.adjust(pval, method="BH"),
         sig_gene=qval < MAX_QVAL & abs(log2FC) >= MIN_L2FC) %>%
  group_by(gene_id) %>%
  summarize(sig_gene=any(sig_gene))
```


# Analysis
## All Genes
```{r tbl_all_genes}
df_sig_all <- full_join(df_sig_dtu, df_sig_dge, by="gene_id") %>%
  replace_na(list(sig_isoform=FALSE, sig_gene=FALSE))

df_sig_all %$%
  table(sig_gene, sig_isoform)
```

## Multi-UTR Genes
```{r tbl_multi_genes}
df_sig_multi <- left_join(df_sig_dtu, df_sig_dge, by="gene_id") %>%
  replace_na(list(sig_isoform=FALSE, sig_gene=FALSE))

df_sig_multi %$%
  table(sig_gene, sig_isoform) %T>%
  print() %>%
  chisq.test()
```


# Pairwise Ery vs HSC only
## All Genes
```{r}
df_hsc_ery <- full_join(
  {
    df_dtu %>%
      filter(start == "14/Stem", end == "12/Ery") %>%
      dplyr::rename(dLUI=stat, pval_dLUI=pval) %>%
      dplyr::select(gene_id, dLUI, pval_dLUI) %>%
      mutate(qval_dLUI=p.adjust(pval_dLUI, method="BH"),
             sig_isoform=qval_dLUI < MAX_QVAL & abs(dLUI) >= MIN_DLUI)}, 
  {
    df_dge %>%
      filter(start == "14/Stem", end == "12/Ery") %>%
      dplyr::rename(pval_L2FC=pval) %>%
      dplyr::select(gene_id, log2FC, pval_L2FC) %>%
      mutate(qval_L2FC=p.adjust(pval_L2FC, method="BH"),
             sig_gene=qval_L2FC < MAX_QVAL & abs(log2FC) >= MIN_L2FC)
  }, by="gene_id") %>%
  replace_na(list(sig_isoform=FALSE, sig_gene=FALSE))

df_hsc_ery %$%
  table(sig_gene, sig_isoform)
```

## Coexpressed Multi-UTR Genes
```{r}
df_hsc_ery %>%
  filter(!is.na(dLUI)) %$%
  table(sig_gene, sig_isoform) %T>%
  print() %>%
  chisq.test()
```

## Plot
```{r plot_l2fc_dlui, fig.width=4, fig.height=4}
df_hsc_ery %>%
  filter(!is.na(log2FC), !is.na(dLUI)) %>%
  ggplot(aes(x=dLUI, y=log2FC)) +
  geom_point(size=1) +
  geom_vline(xintercept=c(-MIN_DLUI, MIN_DLUI), linetype="dashed") +
  geom_hline(yintercept=c(-MIN_L2FC, MIN_L2FC), linetype="dashed") +
  theme_bw()
```


---

# Runtime Details
## Session Info
```{r sesh_info, echo=FALSE}
sessionInfo()
```

## Conda Environment
```{bash comment="", echo=FALSE}
if ! command -v conda &> /dev/null
then
  echo "Conda not detected."
elif [ -z "${CONDA_PREFIX}" ]
then
  echo "No active Conda environment."
else
  echo "## Conda Environment YAML"
  conda env export
fi
```
