---
title: "Two-Sample Test - HSC-Ery"
author: "Mervin M Fansler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    code_folding: show
    toc: true
    toc_float: true
---

# Purpose

This document demonstrates the use of a two-sample bootstrap test using the 
HSPC data from *Dahlin et al., 2018* and *Wolf et al., 2019*.

# Initialization

## Libraries
```{r libs, message=FALSE, warning=FALSE}
library(magrittr)
library(tidyverse)
library(cowplot)
library(ggrastr)
```

## Parameters
```{r set_params}
set.seed(20211017)
MIN_DIPA=0.15
MIN_DLUI=0.15
MIN_WD=0.15
MAX_QVAL=0.05
```

## Functions

```{r methods}
luiVolcanoPlot <- function (df, max_q=0.05, min_dlui=0.1, max_dlui=0.5, legend=TRUE) {
  df %<>% 
    mutate(sig=ifelse(qval < max_q, sprintf("q < %0.2f", max_q), "NS")) %>%
    mutate(size_effect=ifelse(abs(stat) < min_dlui, 
                              sprintf("|dLUI| < %0.2f", min_dlui),
                              ifelse(stat < 0, 
                              sprintf("dLUI < -%0.2f", min_dlui),
                              sprintf("dLUI > %0.2f", min_dlui)))) %>%
    mutate(sig_size=paste(size_effect, sig, sep=", ")) %>%
    mutate(outlier=abs(stat) > max_dlui,
           stat_adj=ifelse(outlier, sign(stat)*max_dlui, stat))
  
  pmain <- df %>% 
    ggplot(aes(x=stat_adj, y=-log10(pval))) +
    rasterize(geom_point(aes(color=sig_size, shape=outlier, size=sig_size)), dpi=300) +
    geom_vline(xintercept=min_dlui*c(-1,1), 
               linetype='dashed', size=0.5, color="darkgrey") +
    scale_color_manual(values=c("grey", "#958DBE", "darkgrey", 
                                "#3F6CA5", "darkgrey", "#C63732")) +
    scale_x_continuous(breaks=scales::pretty_breaks(6), 
                       limits=max_dlui*c(-1,1)) +
    scale_shape_manual(values=c("FALSE"=16, "TRUE"=17)) +
    scale_size_manual(values=c(0.3, 0.3, 0.3, 1.2, 0.3, 1.2)) +
    labs(x="dLUI", 
         y="-log10(p-value)", color="Significance\n (BH Adjusted)") +
    theme(legend.title.align=0.5) + 
    guides(shape=FALSE) +
    theme_cowplot()
  
  if (!legend) {
    pmain <- pmain + guides(color=FALSE, size=FALSE)
  }
  
  xdens <- axis_canvas(pmain, axis='x') + 
    geom_density(data=df, aes(x=stat, fill=sig), alpha=0.7, size=0.2) +
    scale_fill_manual(values=c("black", "#958DBE"))
  
  insert_xaxis_grob(pmain, xdens, grid::unit(0.2, 'null'), position='top') %>% 
    ggdraw()
}

ipaVolcanoPlot <- function (df, max_q=0.05, min_dipa=0.1, max_dipa=0.5, legend=TRUE) {
  df %<>% 
    mutate(sig=ifelse(qval < max_q, sprintf("q < %0.2f", max_q), "NS")) %>%
    mutate(size_effect=ifelse(abs(stat) > min_dipa, 
                              sprintf("dIPA > %0.2f", min_dipa), 
                              sprintf("dIPA < %0.2f", min_dipa))) %>%
    mutate(sig_size=paste(size_effect, sig, sep=", ")) %>%
    mutate(outlier=abs(stat) > max_dipa,
           stat_adj=ifelse(outlier, sign(stat)*max_dipa, stat))
  
  pmain <- df %>% 
    ggplot(aes(x=stat_adj, y=-log10(pval))) +
    rasterize(geom_point(aes(color=sig_size, shape=outlier), size=0.7), dpi=300) +
    geom_vline(xintercept=min_dipa*c(-1,1), 
               linetype='dashed', size=0.5, color="darkgrey") +
    scale_color_manual(values=c("grey", "#663333", "darkgrey", "red")) +
    scale_shape_manual(values=c("FALSE"=16, "TRUE"=17)) +
    scale_x_continuous(breaks=scales::pretty_breaks(6), 
                       limits=max_dipa*c(-1,1)) +
    labs(x="dIPA", 
         y="-log10(p-value)", color="Significance\n (BH Adjusted)") +
    guides(shape=FALSE) +
    theme(legend.title.align=0.5) + 
    theme_cowplot()
  
  if (!legend) {
    pmain <- pmain + guides(color=FALSE)
  }
  
  xdens <- axis_canvas(pmain, axis='x') + 
    geom_density(data=df, aes(x=stat, fill=sig), alpha=0.7, size=0.2) +
    scale_fill_manual(values=c("black", "red"))
  
  insert_xaxis_grob(pmain, xdens, grid::unit(0.2, 'null'), position='top') %>% 
    ggdraw()
}

wdVolcanoPlot <- function (df, max_q=0.05, min_wd=0.1, max_wd=0.5, legend=TRUE) {
  df %<>% 
    mutate(sig=ifelse(qval < max_q, sprintf("q < %0.2f", max_q), "NS")) %>%
    mutate(size_effect=ifelse(stat > min_wd, 
                              sprintf("WD > %0.2f", min_wd), 
                              sprintf("WD < %0.2f", min_wd))) %>%
    mutate(sig_size=paste(size_effect, sig, sep=", ")) %>%
    mutate(outlier=stat > max_wd,
           stat_adj=ifelse(outlier, max_wd, stat))
  
  pmain <- df %>% 
    ggplot(aes(x=stat_adj, y=-log10(pval))) +
    rasterize(geom_point(aes(color=sig_size), size=0.7), dpi=300) +
    geom_vline(xintercept=min_wd, linetype='dashed', size=0.5, color="darkgrey") +
    scale_color_manual(values=c("grey", "#663333", "darkgrey", "red")) +
    scale_shape_manual(values=c("FALSE"=16, "TRUE"=17)) +
    scale_x_continuous(breaks=scales::pretty_breaks(6), 
                       limits=c(0,max_wd)) +
    labs(x="Total Isoform Change", 
         y="-log10(p-value)", color="Significance\n (BH Adjusted)") +
    guides(shape=FALSE) +
    theme(legend.title.align=0.5) + 
    theme_cowplot()
  
  if (!legend) {
    pmain <- pmain + guides(color=FALSE)
  }
  
  xdens <- axis_canvas(pmain, axis='x') + 
    geom_density(data=df, aes(x=stat, fill=sig), alpha=0.7, size=0.2) +
    scale_fill_manual(values=c("black", "red"))
  
  insert_xaxis_grob(pmain, xdens, grid::unit(0.2, 'null'), position='top') %>% 
    ggdraw()
}
```

# Data
## Loading
```{r load_data, message=FALSE}
df_test_ery_hsc_wd <- read_tsv("data/dtu/20211117-hsc-ery-twosample-wd.min50.boot10k.tsv.gz")
df_test_ery_hsc_lui <- read_tsv("data/dtu/20211117-hsc-ery-twosample-lui.min50.boot10k.tsv.gz")
df_test_ery_hsc_ipa <- read_tsv("data/dtu/20211117-hsc-ery-twosample-ipa.min50.boot10k.tsv.gz")
```

# Analysis

## WD Test: Ery vs HSC
### Significant Genes
```{r sig_wd}
df_test_ery_hsc_wd %>%
  mutate(sig_gene=qval < MAX_QVAL & stat >= MIN_WD) %$%
  table(sig_gene)
```

### Volcano Plot
```{r plot_volcano_wd, fig.width=7, fig.height=5}
wdVolcanoPlot(df_test_ery_hsc_wd, max_q=0.05, min_wd=0.15)
```

```{r plot_volcano_wd_save, fig.width=5, fig.height=5}
wdVolcanoPlot(df_test_ery_hsc_wd, max_q=0.05, min_wd=0.15, legend=FALSE)
ggsave("output/figure4/fig4c-ery-hsc-wd-volcano.pdf", width=5, height=5, dpi=300)
```

## LUI Test: Ery vs HSC
### Significant Genes
```{r sig_lui}
df_test_ery_hsc_lui %>%
  mutate(sig_gene=qval < MAX_QVAL & abs(stat) >= MIN_DLUI,
         direction=sign(stat)) %$%
  table(sig_gene, direction)
```

### Significant Genes, By Effect
```{r sig_lui_effect}
df_test_ery_hsc_lui %>%
  mutate(sig=qval < MAX_QVAL,
         abs_dlui=ifelse(abs(stat) < MIN_DLUI, "< 0.15", ">=0.15")) %$%
  table(sig, abs_dlui)
```

### Volcano Plot
```{r plot_volcano_lui, fig.width=7, fig.height=5}
luiVolcanoPlot(df_test_ery_hsc_lui, max_q=0.05, min_dlui=0.15)
```

```{r plot_volcano_lui_save, fig.width=5, fig.height=5}
luiVolcanoPlot(df_test_ery_hsc_lui, max_q=0.05, min_dlui=0.15, legend=FALSE)
ggsave("output/figure4/fig4c-ery-hsc-lui-volcano.pdf", width=5, height=5, dpi=300)
```


## IPA Test: Ery vs HSC
### Significant Genes
```{r sig_ipa}
df_test_ery_hsc_ipa %>%
  mutate(sig_gene=qval < MAX_QVAL & abs(stat) >= MIN_DIPA,
         direction=sign(stat)) %$%
  table(sig_gene, direction)
```

### Volcano Plot
```{r plot_volcano_ipa, fig.width=7, fig.height=5}
ipaVolcanoPlot(df_test_ery_hsc_ipa, max_q=0.05, min_dipa=0.15)
```

```{r plot_volcano_ipa_save, fig.width=5, fig.height=5}
ipaVolcanoPlot(df_test_ery_hsc_ipa, max_q=0.05, min_dipa=0.15, legend=FALSE)
ggsave("output/figure4/fig4c-ery-hsc-ipa-volcano.pdf", width=5, height=5, dpi=300)
```


---

# Runtime Details
## Session Info
```{r sesh_info, echo=FALSE}
sessionInfo()
```

## Conda Environment
```{bash comment="", echo=FALSE}
if ! command -v conda &> /dev/null
then
  echo "Conda not detected."
elif [ -z "${CONDA_PREFIX}" ]
then
  echo "No active Conda environment."
else
  echo "## Conda Environment YAML"
  conda env export
fi
```
