---
title: "Expanded Cell Types - UTRome Tests - GENCODE vs UTRome"
author: "Mervin M Fansler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    code_folding: show
    toc: true
    toc_float: true
---

# Purpose

This compares the overlap of significant genes detected by differential isoform 
usage testing starting from GENCODE versus UTRome annotation counts.

# Initialization

## Libraries
```{r libs, message=FALSE, warning=FALSE}
library(magrittr)
library(tidyverse)
library(writexl)
library(cowplot)
```

## Parameters
```{r set_params}
MAX_QVAL=0.05
MIN_DWUI=0.10

TEST_UTROME="data/dtu/20230523-diff-pairwise-wui-tests.min50.boot10k.tsv.gz"
TEST_GENCODE="data/dtu/20230523-diff-pairwise-wui-tests-gc.min50.boot10k.tsv.gz"
```

# Data
## Loading
```{r load_data, message=FALSE}
df_utrome <- read_tsv(TEST_UTROME) %>% dplyr::rename(gene_id=gene)
df_gencode <- read_tsv(TEST_GENCODE) %>% dplyr::rename(gene_id=gene)
```

## Preprocessing
```{r prepare_data}
df_sig_utrome <- df_utrome %>%
  mutate(sig_utrome=qval < MAX_QVAL & abs(stat) >= MIN_DWUI) %>%
  select(gene_id, start, end, sig_utrome)

df_sig_gencode <- df_gencode %>%
  mutate(sig_gencode=qval < MAX_QVAL & abs(stat) >= MIN_DWUI) %>%
  select(gene_id, start, end, sig_gencode)

df_sig_all <- full_join(df_sig_utrome, df_sig_gencode, by=c("gene_id", "start", "end")) %>%
  replace_na(list(sig_utrome=FALSE, sig_gencode=FALSE)) %>%
  mutate(sig=ifelse(sig_utrome & sig_gencode, "Both", 
                    ifelse(sig_utrome, "UTRome", 
                           ifelse(sig_gencode, "GENCODE", "NS"))),
         sig=factor(sig, levels=c("NS", "UTRome", "Both", "GENCODE")),
         test_label=str_c(end, " vs ", start))
```

### Summary Table
```{r tbl_summary}
df_summary_utrome <- df_utrome %>% 
  group_by(start, end) %>% 
  summarise(n_tests=dplyr::n(), 
            n_sig=sum(abs(stat) >= MIN_DWUI & qval < MAX_QVAL), 
            frac_sig=n_sig/n_tests,
            .groups='drop') 

df_summary_gencode <- df_gencode %>% 
  group_by(start, end) %>% 
  summarise(n_tests=dplyr::n(), 
            n_sig=sum(abs(stat) >= MIN_DWUI & qval < MAX_QVAL), 
            frac_sig=n_sig/n_tests,
            .groups='drop')

df_summary <- full_join(df_summary_utrome, df_summary_gencode,
                        by=c("start", "end"), suffix=c("_UTRome", "_GENCODE"))
```

```{r export_summary}
write_xlsx(df_summary, "tables/som-utrome-vs-gencode-wui-tests.dwui10.fdr5.xlsx")
```

# Summary Plots
```{r fig.width=3, fig.height=3}
df_summary %>%
  ggplot(aes(x=frac_sig_GENCODE, y=frac_sig_UTRome)) +
  geom_point(size=1) +
  geom_abline(slope=1, linewidth=0.1, linetype='dashed', color='orchid') +
  coord_equal(xlim=c(0,0.35), ylim=c(0,0.35)) +
  labs(x="Fraction Significant [GENCODE]",
       y="Fraction Significant [UTRome]") +
  theme_bw()

df_summary %>%
  ggplot(aes(x=n_tests_GENCODE, y=n_tests_UTRome)) +
  geom_point(size=1) +
  geom_abline(slope=1, linewidth=0.1, linetype='dashed', color='orchid') +
  coord_equal(xlim=c(0,5e3), ylim=c(0,5e3)) +
  labs(x="Testable Genes [GENCODE]",
       y="Testable Genes [UTRome]") +
  theme_bw()

df_summary %>%
  ggplot(aes(x=n_sig_GENCODE, y=n_sig_UTRome)) +
  geom_point(size=1) +
  geom_abline(slope=1, linewidth=0.1, linetype='dashed', color='orchid') +
  coord_equal(xlim=c(0,1100), ylim=c(0,1100)) +
  labs(x="Significant Genes [GENCODE]",
       y="Significant Genes [UTRome]") +
  theme_bw()
```

```{r plot_fraction, fig.width=6, fig.height=4}
test_lbl_lvls <- c("MEF vs mESC",
                   "ODC vs mNeur",
                   "MG vs mNeur",
                   "Ery vs HSC",
                   "endo vs mNeur",
                   "Astro vs mNeur",
                   "ODC vs OPC",
                   "Mono vs HSC",
                   "Neutro vs HSC",
                   "TraEpi vs TraMes",
                   "TSpleen vs Timm",
                   "basMamm vs lumMamm",
                   "TMamm vs TSpleen", 
                   "mNeur vs imNeur",
                   "Baso vs HSC",
                   "BSpleen vs TSpleen",
                   "NeurEC vs mNeur")

df_summary_long_frac <- df_summary %>%
  mutate(test_label=fct(str_c(end, " vs ", start), test_lbl_lvls)) %>%
  dplyr::select(test_label, frac_sig_GENCODE, frac_sig_UTRome) %>%
  pivot_longer(cols=starts_with("frac_sig"), 
               names_to="target", values_to="frac_sig",
               names_prefix="frac_sig_") %>%
  mutate(target=fct(target, c("GENCODE", "UTRome")))

df_summary_long_frac %>%
  arrange(desc(target)) %>%
  ggplot(aes(x=test_label, y=frac_sig, fill=target)) +
  geom_bar(stat='identity', position=position_identity(), 
           color='black', linewidth=0.1) +
  scale_fill_manual(values=c("UTRome"='steelblue', "GENCODE"='grey80')) +
  scale_x_discrete(guide=guide_axis(angle=45)) +
  scale_y_continuous(expand=c(0,0,0.05,0)) +
  labs(x=NULL, y="Fraction of Significant DTU Tests",
       fill="scUTRquant\nTarget") +
  theme_bw()

df_summary_long_frac %>%
  ggplot(aes(x=test_label, y=frac_sig, fill=target)) +
  geom_bar(stat='identity', position='dodge',
           color='black', linewidth=0.1, width=0.8) +
  scale_fill_manual(values=c("GENCODE"='grey80', "UTRome"='steelblue')) +
  scale_x_discrete(guide=guide_axis(angle=45)) +
  scale_y_continuous(expand=c(0,0,0.05,0)) +
  labs(x=NULL, y="Fraction of Significant DTU Tests",
       fill="scUTRquant\nTarget") +
  theme_bw()

ggsave("output/figure4/fig4f-celltypes-utrome-gc-fraction-dodged.pdf", width=6, height=4, dpi=300)
```


### Absolute Numbers
```{r abs_dodged, width=6, height=4}
df_summary_long_n <- df_summary %>%
  mutate(test_label=fct(str_c(end, " vs ", start), test_lbl_lvls)) %>%
  dplyr::select(test_label, n_sig_GENCODE, n_sig_UTRome) %>%
  pivot_longer(cols=starts_with("n_sig"), 
               names_to="target", values_to="n_sig",
               names_prefix="n_sig_") %>%
  mutate(target=fct(target, c("GENCODE", "UTRome")))

df_summary_long_n %>%
  arrange(desc(target)) %>%
  ggplot(aes(x=test_label, y=n_sig, fill=target)) +
  geom_bar(stat='identity', position=position_identity(), 
           color='black', linewidth=0.1) +
  scale_fill_manual(values=c("UTRome"='steelblue', "GENCODE"='grey80')) +
  scale_x_discrete(guide=guide_axis(angle=45)) +
  scale_y_continuous(expand=c(0,0,0.05,0)) +
  labs(x=NULL, y="Significant DTU Tests",
       fill="scUTRquant\nTarget") +
  theme_bw()

df_summary_long_n %>%
  ggplot(aes(x=test_label, y=n_sig, fill=target)) +
  geom_bar(stat='identity', position='dodge',
           color='black', linewidth=0.1, width=0.8) +
  scale_fill_manual(values=c("GENCODE"='grey80', "UTRome"='steelblue')) +
  scale_x_discrete(guide=guide_axis(angle=45)) +
  scale_y_continuous(expand=c(0,0,0.05,0)) +
  labs(x=NULL, y="Significant DTU Tests",
       fill="scUTRquant\nTarget") +
  theme_bw()

ggsave("output/figure4/fig4f-celltypes-utrome-gc-nsig-dodged.pdf", width=6, height=4, dpi=300)
```


# All Genes
## Everything
```{r plot_all_ns, fig.width=6, fig.height=5}
colvals_ns <- c(NS="#CCCCCC", UTRome="#EEE643", Both="#2E9F72", GENCODE="#2170B1") %>% {
  setNames(colorspace::desaturate(., 0.2), names(.))
}

df_sig_all %>%
  mutate(test_label=fct_reorder(test_label, sig, . %>% { sum(. != "NS") }, .desc=TRUE)) %>%
  ggplot(aes(x=test_label, fill=sig)) +
  geom_bar(position='stack', color='black', linewidth=0.1) +
  scale_fill_manual(breaks=names(colvals_ns), values=colvals_ns) +
  scale_x_discrete(guide=guide_axis(angle=45)) +
  scale_y_continuous(expand=c(0,0,0.05,0)) +
  labs(x="Compared Cell Types", y="All Tested Genes", fill="Difference") +
  theme_bw()

ggsave("output/figure4/fig4f-celltypes-utrome-gc-compared-all-ns.pdf", width=6, height=5, dpi=300)
```

## Plot Positive
```{r plot_all, fig.width=6, fig.height=5}
colvals <- c(UTRome="#EEE643", Both="#2E9F72", GENCODE="#2170B1") %>% {
  setNames(colorspace::desaturate(., 0.2), names(.))
}

df_sig_all %>%
  filter(sig != "NS") %>%
  ggplot(aes(x=fct_infreq(test_label), fill=sig)) +
  geom_bar(position='stack', color='black', linewidth=0.1) +
  scale_fill_manual(values=colvals) +
  scale_x_discrete(guide=guide_axis(angle=45)) +
  scale_y_continuous(expand=c(0,0,0.05,0)) +
  labs(x="Compared Cell Types", y="Significant Genes", fill="Difference") +
  theme_bw()

ggsave("output/figure4/fig4f-celltypes-utrome-gc-compared-all.pdf", width=6, height=5, dpi=300)
```

```{r}
df_ntests_compare <- full_join(df_summary_gencode, df_summary_utrome, by=c("start", "end"), 
          suffix=c("_gencode", "_utrome")) %>%
  mutate(ratio_tests=n_tests_utrome/n_tests_gencode,
         ratio_frac=frac_sig_utrome/frac_sig_gencode)

df_ntests_compare

summary(df_ntests_compare$ratio_tests - 1)
summary(df_ntests_compare$frac_sig_utrome/df_ntests_compare$frac_sig_gencode - 1)
```


```{r tbl_wide}
df_chisq <- df_sig_all %>%
  group_by(start, end) %>%
  nest() %>%
  transmute(start, end, res=map(data, ~ broom::tidy(chisq.test(.x$sig_utrome, .x$sig_gencode, simulate.p.value=TRUE, B=10000)))) %>%
  unnest(res) %>%
  select(start, end, statistic, p.value)

map_sig <- c("FALSE_FALSE"="NS",
             "FALSE_TRUE"="UTRome",
             "TRUE_FALSE"="GENCODE",
             "TRUE_TRUE"="Both")

df_wide <- df_sig_all %>%
  group_by(start, end) %>%
  dplyr::count(sig_gencode, sig_utrome) %>%
  mutate(result=map_sig[str_c(sig_gencode,sig_utrome, sep="_")]) %>%
  ungroup() %>%
  pivot_wider(id_cols=c("start", "end"), names_from="result", values_from="n", values_fill=0L) %>%
  mutate("Coexpressed multi-UTR genes"=NS+UTRome+GENCODE+Both,
         "Fraction more significant"=(UTRome+Both)/(GENCODE+Both)) %>%
  left_join(df_chisq, by=c("start", "end")) %>%
  dplyr::rename("Cell Type 1"=start,
                "Cell Type 2"=end,
                "Chi-squared"=statistic,
                "p-value"=p.value) %>%
  select(`Cell Type 1`, `Cell Type 2`, `Coexpressed multi-UTR genes`,
         NS:Both, `Fraction more significant`, `Chi-squared`, `p-value`)

knitr::kable(df_wide)

summary(df_wide$`Fraction more significant` - 1)
```

## Export Table
```{r export_tbl}
write_xlsx(df_wide, "output/tables/tbl3-extended-celltypes-chisq-results-utrome-gc.xlsx")
```



---

# Runtime Details
## Session Info
```{r sesh_info, echo=FALSE}
sessionInfo()
```

## Conda Environment
```{bash comment="", echo=FALSE}
if ! command -v conda &> /dev/null
then
  echo "Conda not detected."
elif [ -z "${CONDA_PREFIX}" ]
then
  echo "No active Conda environment."
else
  echo "## Conda Environment YAML"
  conda env export
fi
```
