---
title: "LUI Test Results - Erythrocyte Trajectory"
author: "Mervin M Fansler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    code_folding: show
    toc: true
    toc_float: true
---

# Purpose

This analysis processes the two-sample LUI test results to identify significant
isoform changes in the HSC to Erythrocyte trajectory from the *Dahlin et al., 2018*
dataset. Significant genes are then plotted as a heat map.

# Initialization

## Libraries
```{r libs, message=FALSE, warning=FALSE}
library(magrittr)
library(tidyverse)
library(cowplot)
library(SingleCellExperiment)
library(Matrix)
library(matrixStats)
library(pheatmap)
```

## Parameters
```{r set_params}
set.seed(20211020)
MIN_DLUI=0.15
MAX_QVAL=0.05
MIN_NCELLS=50
MIN_FC=1.5
MIN_L2FC=log2(MIN_FC)

TSV_DGE_TESTS="data/dge/20221222-ery-trajectory-pairwise-t-tests.tsv.gz"
TSV_LUI_TESTS="data/dtu/20211019-ery-trajectory-pairwise-lui-tests.min50.boot10k.tsv.gz"
```

## Functions

```{r methods}

```

# Data
## Loading
```{r load_data, message=FALSE}
df_tests <- read_tsv(TSV_LUI_TESTS, 
                     col_types='ccccddi') %>%
  mutate(test=str_c(end, " vs ", start),
         qval=p.adjust(pval, method='BH'),
         sig=(qval < MAX_QVAL) & (abs(stat) >= MIN_DLUI))

df_dge <- read_tsv(TSV_DGE_TESTS, col_types='ccccddd') %>%
  mutate(test=str_c(end, " vs ", start),
         qval=p.adjust(pval, method='BH'),
         sig=(qval < MAX_QVAL) & (abs(log2FC) >= MIN_L2FC))

sce <- readRDS("data/sce/hspcs.txs.full_annot.Rds")
```

## Preprocessing
### Compute Summary Tables
```{r compute_summaries}
df_summary <- df_tests %>%
  group_by(test) %>%
  summarize(n_tested=n(),
            n_shortening=sum(sig & stat < 0, na.rm=TRUE),
            n_lengthening=sum(sig & stat > 0, na.rm=TRUE),
            pct_shortening=100*n_shortening/n_tested,
            pct_lengthening=100*n_lengthening/n_tested,
            .groups='drop')

df_sig_genes <- df_tests %>%
  filter(sig) %>%
  group_by(gene_symbol, gene) %>%
  summarize(max_test=test[which.max(abs(stat))],
            max_dlui=stat[which.max(abs(stat))],
            .groups='drop')

genes_lengthening <- filter(df_sig_genes, max_dlui > 0)$gene
genes_shortening  <- filter(df_sig_genes, max_dlui < 0)$gene

genes_dge <- filter(df_sig_genes, gene %in% unique(filter(df_dge, sig)$gene_id))$gene
genes_lengthening_nodge <- setdiff(genes_lengthening, genes_dge)
genes_shortening_nodge <- setdiff(genes_shortening, genes_dge)
```

### Compute Significant LUIs
```{r compute_luis}
celltypes <- unique(c(df_tests$start, df_tests$end))

idx_cells <- sce$clusters %in% celltypes
idx_sig_genes <- rowData(sce) %>%
  { (.$gene_id %in% df_sig_genes$gene) & (!.$is_ipa) }

## filter sce
sce %<>% `[`(idx_sig_genes, idx_cells)
cts_tx_cell <- assay(sce, "normcounts")

## design matrices
M_cell_celltype <- t(fac2sparse(sce$clusters))[,celltypes]
M_gene_tx <- fac2sparse(rowData(sce)$gene_id)
M_lu_tx <- M_gene_tx %*% Diagonal(nrow(sce), rowData(sce)$is_distal)

## compute luis
cts_tx_celltype <- cts_tx_cell %*% M_cell_celltype
lui_gene_celltype <- (M_lu_tx %*% cts_tx_celltype) / (M_gene_tx %*% cts_tx_celltype)

## count expressing cells
ncells_gene_celltype <- ((M_gene_tx %*% cts_tx_cell) > 0) %*% M_cell_celltype

## filter luis
idx_expressed_genes <- rowAlls(drop0(ncells_gene_celltype >= MIN_NCELLS), value=TRUE)
flui_gene_celltype <- as.matrix(lui_gene_celltype[idx_expressed_genes,])

dlui_gene_celltype <- t(flui_gene_celltype) %>% scale(scale=FALSE) %>% t()
zdlui_gene_celltype <- t(flui_gene_celltype) %>% scale() %>% t()
```


# Summary Overview
## Significant Tests

```{r sig_tests}
knitr::kable(df_summary)
```

## Significant Genes
```{r sig_genes}
df_sig_genes %>%
  summarize(n_shortening=sum(max_dlui < 0),
            n_lengthening=sum(max_dlui > 0)) %>%
  knitr::kable()

df_sig_genes %>%
  group_by(max_test) %>%
  summarize(n_shortening=sum(max_dlui < 0),
            n_lengthening=sum(max_dlui > 0),
            .groups='drop') %>%
  knitr::kable()
```



# Plot Heatmaps

## LUI
### Longer-Shorter Order
```{r compute_gene_orders1}
lui_medians <- setNames(rowMedians(flui_gene_celltype), rownames(flui_gene_celltype))

idx_lengthening <- which(names(lui_medians) %in% genes_lengthening) %>% `[`(order(lui_medians[.]))
idx_shortening  <- which(names(lui_medians) %in% genes_shortening) %>% `[`(order(lui_medians[.]))

idx_ordered <- c(idx_lengthening, idx_shortening)
```

We observe `r length(idx_lengthening)` genes lengthening and `r length(idx_shortening)` genes shortening.

```{r fig_heatmap_lui, fig.width=4, fig.height=4}
NCOLORS=100
breaks.point <- seq(0, 1, length.out=NCOLORS + 1)

pheatmap(flui_gene_celltype[idx_ordered,], 
         cluster_rows=FALSE, cluster_cols=FALSE, 
         labels_col=c("HSC", "Prog1", "Prog2", "MEP", "ProE", "Baso1", "Baso2", "Ery"),
         show_rownames=FALSE, scale='none', fontsize=8,
         color=colorRampPalette(c("#336BA8", "white", "#D62125"))(NCOLORS),
         breaks=breaks.point, angle_col=0, 
         filename="output/figure4/fig4d-lui-hsc-ery-path.pdf",
         width=4, height=4)
```
### DGE-Longer-Shorter Order
```{r compute_gene_orders2}
idx_dge <- which(names(lui_medians) %in% genes_dge) %>% `[`(order(lui_medians[.]))
idx_lengthening_nodge <- which(names(lui_medians) %in% genes_lengthening_nodge) %>% `[`(order(lui_medians[.]))
idx_shortening_nodge  <- which(names(lui_medians) %in% genes_shortening_nodge) %>% `[`(order(lui_medians[.]))

idx_ordered <- c(idx_dge, idx_lengthening_nodge, idx_shortening_nodge)
```

We observe `r length(idx_dge)` genes with DGE and dLUI, `r length(idx_lengthening_nodge)` genes lengthening only, and `r length(idx_shortening_nodge)` genes shortening only.

```{r fig_heatmap_lui_sepdge, fig.width=4, fig.height=4}
NCOLORS=100
breaks.point <- seq(0, 1, length.out=NCOLORS + 1)

pheatmap(flui_gene_celltype[idx_ordered,], 
         cluster_rows=FALSE, cluster_cols=FALSE, 
         labels_col=c("HSC", "Prog1", "Prog2", "MEP", "ProE", "Baso1", "Baso2", "Ery"),
         show_rownames=FALSE, scale='none', fontsize=8,
         color=colorRampPalette(c("#336BA8", "white", "#D62125"))(NCOLORS),
         breaks=breaks.point, angle_col=0,
         filename="output/figure4/fig4d-lui-hsc-ery-path-sepdge.pdf",
         width=4, height=4)
```

## DLUI
### Longer-Shorter Order
```{r compute_gene_orders3}
wt_dlui <- (dlui_gene_celltype %*% seq(8))[,1]

idx_lengthening <- which(names(wt_dlui) %in% genes_lengthening) %>% `[`(order(wt_dlui[.]))
idx_shortening  <- which(names(wt_dlui) %in% genes_shortening) %>% `[`(order(wt_dlui[.]))

idx_ordered <- c(idx_lengthening, idx_shortening)
```

We observe `r length(idx_lengthening)` genes lengthening and `r length(idx_shortening)` genes shortening.

```{r fig_heatmap_dlui, fig.width=4, fig.height=4}
NCOLORS=100
breaks.point <- seq(-0.3, 0.3, length.out=NCOLORS + 1)

pheatmap(dlui_gene_celltype[idx_ordered,], 
         cluster_rows=FALSE, cluster_cols=FALSE, 
         labels_col=c("HSC", "Prog1", "Prog2", "MEP", "ProE", "Baso1", "Baso2", "Ery"),
         show_rownames=FALSE, scale='none', fontsize=8,
         color=colorRampPalette(c("#336BA8", "white", "#D62125"))(NCOLORS),
         breaks=breaks.point, angle_col=0,
         filename="output/figure4/fig4d-dlui-hsc-ery-path.pdf",
         width=4, height=4)
```

### DGE-Longer-Shorter Order
```{r compute_gene_orders4}
idx_dge <- which(names(wt_dlui) %in% genes_dge) %>% `[`(order(wt_dlui[.]))
idx_lengthening_nodge <- which(names(wt_dlui) %in% genes_lengthening_nodge) %>% `[`(order(wt_dlui[.]))
idx_shortening_nodge  <- which(names(wt_dlui) %in% genes_shortening_nodge) %>% `[`(order(wt_dlui[.]))

idx_ordered <- c(idx_dge, idx_lengthening_nodge, idx_shortening_nodge)
```

We observe `r length(idx_dge)` genes with DGE and dLUI, `r length(idx_lengthening_nodge)` genes lengthening only, and `r length(idx_shortening_nodge)` genes shortening only.

```{r fig_heatmap_dlui_sepdge, fig.width=4, fig.height=4}
NCOLORS=100
breaks.point <- seq(-0.3, 0.3, length.out=NCOLORS + 1)

pheatmap(dlui_gene_celltype[idx_ordered,], 
         cluster_rows=FALSE, cluster_cols=FALSE, 
         labels_col=c("HSC", "Prog1", "Prog2", "MEP", "ProE", "Baso1", "Baso2", "Ery"),
         show_rownames=FALSE, scale='none', fontsize=8,
         color=colorRampPalette(c("#336BA8", "white", "#D62125"))(NCOLORS),
         breaks=breaks.point, angle_col=0,
         filename="output/figure4/fig4d-dlui-hsc-ery-path-sepdge.pdf",
         width=4, height=4)
```

## ZDLUI
### Longer-Shorter Order
```{r compute_gene_orders5}
wt_zdlui <- (zdlui_gene_celltype %*% seq(8))[,1]

idx_lengthening <- which(names(wt_zdlui) %in% genes_lengthening) %>% `[`(order(wt_zdlui[.]))
idx_shortening  <- which(names(wt_zdlui) %in% genes_shortening) %>% `[`(order(wt_zdlui[.]))

idx_ordered <- c(idx_lengthening, idx_shortening)
```

```{r fig_heatmap_zdlui, fig.width=4, fig.height=4}
NCOLORS=100
breaks.point <- seq(-3, 3, length.out=NCOLORS + 1)

pheatmap(zdlui_gene_celltype[idx_ordered,], 
         cluster_rows=FALSE, cluster_cols=FALSE, 
         labels_col=c("HSC", "Prog1", "Prog2", "MEP", "ProE", "Baso1", "Baso2", "Ery"),
         show_rownames=FALSE, scale='none', fontsize=8,
         color=colorRampPalette(c("#336BA8", "white", "#D62125"))(NCOLORS),
         breaks=breaks.point, angle_col=0,
         filename="output/figure4/fig4d-zdlui-hsc-ery-path.pdf",
         width=4, height=4)
```

### DGE-Longer-Shorter Order
```{r compute_gene_orders6}
idx_dge <- which(names(wt_zdlui) %in% genes_dge) %>% `[`(order(wt_zdlui[.]))
idx_lengthening_nodge <- which(names(wt_zdlui) %in% genes_lengthening_nodge) %>% `[`(order(wt_zdlui[.]))
idx_shortening_nodge  <- which(names(wt_zdlui) %in% genes_shortening_nodge) %>% `[`(order(wt_zdlui[.]))

idx_ordered <- c(idx_dge, idx_lengthening_nodge, idx_shortening_nodge)
```


```{r fig_heatmap_zdlui_sepdge, fig.width=4, fig.height=4}
NCOLORS=100
breaks.point <- seq(-3, 3, length.out=NCOLORS + 1)

pheatmap(zdlui_gene_celltype[idx_ordered,], 
         cluster_rows=FALSE, cluster_cols=FALSE, 
         labels_col=c("HSC", "Prog1", "Prog2", "MEP", "ProE", "Baso1", "Baso2", "Ery"),
         show_rownames=FALSE, scale='none', fontsize=8,
         color=colorRampPalette(c("#336BA8", "white", "#D62125"))(NCOLORS),
         breaks=breaks.point, angle_col=0,
         filename="output/figure4/fig4d-zdlui-hsc-ery-path-sepdge.pdf",
         width=4, height=4)
```

---

# Runtime Details
## Session Info
```{r sesh_info, echo=FALSE}
sessionInfo()
```

## Conda Environment
```{bash comment="", echo=FALSE}
if ! command -v conda &> /dev/null
then
  echo "Conda not detected."
elif [ -z "${CONDA_PREFIX}" ]
then
  echo "No active Conda environment."
else
  echo "## Conda Environment YAML"
  conda env export
fi
```
