---
title: "Expanded Cell Types - DGE vs DTU Tests"
author: "Mervin M Fansler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    code_folding: show
    toc: true
    toc_float: true
---

# Purpose

This compares the overlap of significant genes detected by differential gene 
expression testing versus differential isoform usage testing.

# Initialization

## Libraries
```{r libs, message=FALSE, warning=FALSE}
library(magrittr)
library(tidyverse)
library(writexl)
library(cowplot)
```

## Parameters
```{r set_params}
MAX_QVAL=0.05
MIN_L2FC=log2(1.5)
MIN_DWUI=0.10

TEST_DTU="data/dtu/20230523-diff-pairwise-wui-tests.min50.boot10k.tsv.gz"
TEST_DGE="data/dge/20230530-diff-pairwise-t-tests.tsv.gz"
```

# Data
## Loading
```{r load_data, message=FALSE}
df_dtu <- read_tsv(TEST_DTU) %>%
  dplyr::rename(gene_id=gene)
df_dge <- read_tsv(TEST_DGE)
```

## Preprocessing
```{r prepare_data}
gene_multi <- unique(df_dtu$gene_id)

df_sig_dtu <- df_dtu %>%
  mutate(sig_isoform=qval < MAX_QVAL & abs(stat) >= MIN_DWUI) %>%
  select(gene_id, start, end, sig_isoform)

df_sig_dge <- df_dge %>%
    mutate(sig_gene=qval < MAX_QVAL & abs(log2FC) >= MIN_L2FC) %>%
  select(gene_id, start, end, sig_gene)

df_sig_all <- full_join(df_sig_dtu, df_sig_dge, by=c("gene_id", "start", "end")) %>%
  replace_na(list(sig_isoform=FALSE, sig_gene=FALSE)) %>%
  mutate(sig=ifelse(sig_isoform & sig_gene, "Both", 
                    ifelse(sig_isoform, "DTU", 
                           ifelse(sig_gene, "DGE", "NS"))),
         sig=factor(sig, levels=c("NS", "DTU", "Both", "DGE")),
         test_label=str_c(end, " vs ", start))

df_sig_split <- df_sig_all %>%
  filter(sig != "NS") %>%
  mutate(is_single=!gene_id %in% gene_multi,
         sig=as.character(sig),
         sig=ifelse(sig != "DGE", sig,
                    ifelse(is_single, "DGE (single)", "DGE (multi)")),
         sig=factor(sig, levels=c("DTU", "Both", "DGE (multi)", "DGE (single)")))

df_sig_multi <- left_join(df_sig_dtu, df_sig_dge, by=c("gene_id", "start", "end")) %>%
  replace_na(list(sig_isoform=FALSE, sig_gene=FALSE)) %>%
  mutate(sig=ifelse(sig_isoform & sig_gene, "Both", 
                    ifelse(sig_isoform, "DTU",
                           ifelse(sig_gene, "DGE", "NS"))),
         sig=factor(sig, levels=c("NS", "DTU", "Both", "DGE")),
         test_label=str_c(end, " vs ", start))
```

# All Genes
## Everything
```{r plot_all_ns, fig.width=6, fig.height=5}
colvals_ns <- c(NS="#CCCCCC", DTU="#EEE643", Both="#2E9F72", DGE="#2170B1") %>% {
  setNames(colorspace::desaturate(., 0.2), names(.))
}

df_sig_all %>%
  mutate(test_label=fct_reorder(test_label, sig, . %>% { sum(. != "NS") }, .desc=TRUE)) %>%
  ggplot(aes(x=test_label, fill=sig)) +
  geom_bar(position='stack', color='black', size=0.1) +
  scale_fill_manual(breaks=names(colvals_ns), values=colvals_ns) +
  scale_x_discrete(guide=guide_axis(angle=45)) +
  scale_y_continuous(expand=c(0,0,0.05,0)) +
  labs(x="Compared Cell Types", y="All Tested Genes", fill="Difference") +
  theme_bw()

ggsave("output/figure4/fig4f-celltypes-dge-dtu-compared-all-ns.pdf", width=6, height=5, dpi=300)
```

## Plot Positive
```{r plot_all, fig.width=6, fig.height=5}
colvals <- c(DTU="#EEE643", Both="#2E9F72", DGE="#2170B1") %>% {
  setNames(colorspace::desaturate(., 0.2), names(.))
}

df_sig_all %>%
  filter(sig != "NS") %>%
  ggplot(aes(x=fct_infreq(test_label), fill=sig)) +
  geom_bar(position='stack', color='black', size=0.1) +
  scale_fill_manual(values=colvals) +
  scale_x_discrete(guide=guide_axis(angle=45)) +
  scale_y_continuous(expand=c(0,0,0.05,0)) +
  labs(x="Compared Cell Types", y="Significant Genes", fill="Difference") +
  theme_bw()

ggsave("output/figure4/fig4f-celltypes-dge-dtu-compared-all.pdf", width=6, height=5, dpi=300)
```

## Plot Positive - Split
```{r plot_split, fig.width=7, fig.height=5}
colvals <- c(DTU="#EEE643", Both="#2E9F72", 
             `DGE (multi)`="#2170B1", `DGE (single)`="#512091") %>% {
  setNames(colorspace::desaturate(., 0.2), names(.))
}

df_sig_split %>%
  ggplot(aes(x=fct_infreq(test_label), fill=sig)) +
  geom_bar(position='stack', color='black', size=0.1) +
  scale_fill_manual(values=colvals) +
  scale_x_discrete(guide=guide_axis(angle=90)) +
  scale_y_continuous(expand=c(0,0,0.05,0)) +
  labs(x="Compared Cell Types", y="Significant Genes", fill="Difference") +
  theme_bw()

ggsave("output/figure4/fig4f-celltypes-dge-dtu-compared-split.pdf", width=7, height=5, dpi=300)
```

## Table 
```{r tbl_all_genes}
map_sig <- c("FALSE_FALSE"="NS",
             "FALSE_TRUE"="DTU",
             "TRUE_FALSE"="DGE",
             "TRUE_TRUE"="Both")

df_wide_all <- df_sig_all %>%
  group_by(start, end) %>%
  dplyr::count(sig_gene, sig_isoform) %>%
  mutate(result=map_sig[str_c(sig_gene,sig_isoform, sep="_")]) %>%
  ungroup() %>%
  pivot_wider(id_cols=c("start", "end"), names_from="result", values_from="n", values_fill=0L) %>%
  mutate("All genes"=NS+DTU+DGE+Both) %>%
  dplyr::rename("Cell Type 1"=start, "Cell Type 2"=end) %>%
  select(`Cell Type 1`, `Cell Type 2`, NS:Both)

knitr::kable(df_wide_all)
```

## Export Table
```{r export_tbl_all}
write_xlsx(df_wide_all, "output/tables/sup-tbl3-extended-celltypes-chisq-results-t-tests-all.xlsx")
```

# Multi-UTR Genes
## Plot Positive-Only
```{r plot_multi, fig.width=6, fig.height=5}
colvals <- c(DTU="#EEE643", Both="#2E9F72", DGE="#2170B1") %>% {
  setNames(colorspace::desaturate(., 0.2), names(.))
}

df_sig_multi %>%
  filter(sig != "NS") %>%
  ggplot(aes(x=fct_infreq(test_label), fill=sig)) +
  geom_bar(position='stack', color='black', size=0.1) +
  scale_fill_manual(breaks=names(colvals), values=colvals) +
  scale_x_discrete(guide=guide_axis(angle=45)) +
  scale_y_continuous(expand=c(0,0,0.05,0)) +
  labs(x="Compared Cell Types", y="Significant Multi-UTR Genes", 
       fill="Difference") +
  theme_bw()

ggsave("output/figure4/fig4f-celltypes-dge-dtu-compared-multi.pdf", width=6, height=5, dpi=300)
```

## Plot All
```{r plot_multi_ns, fig.width=6, fig.height=5}
colvals_ns <- c(NS="#CCCCCC", DTU="#EEE643", Both="#2E9F72", DGE="#2170B1") %>% {
  setNames(colorspace::desaturate(., 0.2), names(.))
}

df_sig_multi %>%
  mutate(test_label=fct_reorder(test_label, sig, . %>% { sum(.=="DTU") }, .desc=TRUE)) %>%
  ggplot(aes(x=test_label, fill=sig)) +
  geom_bar(position='stack', color='black', size=0.1) +
  scale_fill_manual(breaks=names(colvals_ns), values=colvals_ns) +
  scale_x_discrete(guide=guide_axis(angle=45)) +
  scale_y_continuous(expand=c(0,0,0.05,0)) +
  labs(x="Compared Cell Types", y="Coexpressed Multi-UTR Genes", 
       fill="Difference") +
  theme_bw()

ggsave("output/figure4/fig4f-celltypes-dge-dtu-compared-multi-ns.pdf", width=6, height=5, dpi=300)
```

## Table 
```{r tbl_multi_genes}
df_sig_multi %>%
  group_by(start, end) %>%
  dplyr::count(sig_gene, sig_isoform) %>%
  knitr::kable()
```

## Chi-Square Tests
```{r}
df_sig_multi %>%
  group_by(start, end) %>%
  nest() %>%
  transmute(start, end, res=map(data, ~ broom::tidy(chisq.test(.x$sig_gene, .x$sig_isoform, simulate.p.value=TRUE, B=10000)))) %>%
  unnest(res) %>%
  select(start, end, statistic, p.value) %>%
  knitr::kable()
```


```{r tbl_wide}
df_chisq <- df_sig_multi %>%
  group_by(start, end) %>%
  nest() %>%
  transmute(start, end, res=map(data, ~ broom::tidy(chisq.test(.x$sig_gene, .x$sig_isoform, simulate.p.value=TRUE, B=10000)))) %>%
  unnest(res) %>%
  select(start, end, statistic, p.value)

map_sig <- c("FALSE_FALSE"="NS",
             "FALSE_TRUE"="DTU",
             "TRUE_FALSE"="DGE",
             "TRUE_TRUE"="Both")

df_wide <- df_sig_multi %>%
  group_by(start, end) %>%
  dplyr::count(sig_gene, sig_isoform) %>%
  mutate(result=map_sig[str_c(sig_gene,sig_isoform, sep="_")]) %>%
  ungroup() %>%
  pivot_wider(id_cols=c("start", "end"), names_from="result", values_from="n", values_fill=0L) %>%
  mutate("Coexpressed multi-UTR genes"=NS+DTU+DGE+Both) %>%
  left_join(df_chisq, by=c("start", "end")) %>%
  dplyr::rename("Cell Type 1"=start,
                "Cell Type 2"=end,
                "Chi-squared"=statistic,
                "p-value"=p.value) %>%
  select(`Cell Type 1`, `Cell Type 2`, `Coexpressed multi-UTR genes`,
         NS:Both, `Chi-squared`, `p-value`)

knitr::kable(df_wide)
```

## Export Table
```{r export_tbl}
write_xlsx(df_wide, "output/tables/tbl3-extended-celltypes-chisq-results-t-tests.xlsx")
```

# Correlation Analysis
```{r plot_correlations, fig.width=6, fig.height=3}
test_lbl_lvls <- c("MEF vs mESC",
                   "ODC vs mNeur",
                   "MG vs mNeur",
                   "Ery vs HSC",
                   "endo vs mNeur",
                   "Astro vs mNeur",
                   "ODC vs OPC",
                   "Mono vs HSC",
                   "Neutro vs HSC",
                   "TraEpi vs TraMes",
                   "TSpleen vs Timm",
                   "basMamm vs lumMamm",
                   "TMamm vs TSpleen", 
                   "mNeur vs imNeur",
                   "Baso vs HSC",
                   "BSpleen vs TSpleen",
                   "NeurEC vs mNeur")

df_corrs <- left_join(
  mutate(df_dtu, test_label=str_c(end, " vs ", start), start=NULL, end=NULL),
  mutate(df_dge, test_label=str_c(end, " vs ", start), start=NULL, end=NULL),
  by=c("gene_symbol", "gene_id", "test_label"), suffix=c("_dtu", "_dge")) %>%
  mutate(test_label=fct(test_label, levels=test_lbl_lvls)) %>%
  group_by(test_label) %>%
  group_modify(~ broom::tidy(cor.test(.x$stat, .x$log2FC, method="pearson")))

df_corrs %>%
  ggplot(aes(x=test_label)) +
  geom_hline(yintercept=0, linetype='dashed', linewidth=0.2) +
  geom_errorbar(aes(ymin=conf.low, ymax=conf.high), 
                linewidth=0.5, width=0.5, color='grey60') +
  geom_point(aes(y=estimate), size=2, color='grey40') +
  scale_x_discrete(guide=guide_axis(angle=90)) +
  scale_y_continuous(limits=c(-0.3,0.3)) +
  labs(x=NULL, y="Pearson Correlation\nlog2(FC) ~ dWUI") +
  theme_bw()

ggsave("output/figure4/sup3f-celltypes-l2fc-dwui-corrs-multi.pdf", width=6, height=3, dpi=300)
```

## Export Table
```{r export_tbl_corrs}
write_xlsx(df_corrs, "output/tables/tbl3b-extended-celltypes-corrs-results.xlsx")
```

---

# Runtime Details
## Session Info
```{r sesh_info, echo=FALSE}
sessionInfo()
```

## Conda Environment
```{bash comment="", echo=FALSE}
if ! command -v conda &> /dev/null
then
  echo "Conda not detected."
elif [ -z "${CONDA_PREFIX}" ]
then
  echo "No active Conda environment."
else
  echo "## Conda Environment YAML"
  conda env export
fi
```
