---
title: "Ery Bootstrap Plots - Cell Type"
author: "Mervin M Fansler"
date: '`r format(Sys.time(), "%d %B %Y")`'
output: 
  html_document: 
    code_folding: show
    toc: true
    toc_float: true
---

# Purpose

Illustrate use of bootstrapping for estimating mean WUI usage. Splitting by batches
to emphasize variance between batches is smaller than variance across celltypes.


# Initialization

## Libraries
```{r libs, message=FALSE, warning=FALSE}
library(SingleCellExperiment)
library(Matrix)
library(matrixStats)
library(magrittr)
library(tidyverse)
library(cowplot)
library(colorspace)
#library(scutrboot)
```

## Parameters
```{r set_params}
set.seed(20211021)
N_BOOTSTRAPS=2000
MIN_NCELLS=50
MIN_UI_RANGE=0.4

genes_to_plot <- c("Sting1", "Ankrd9", "Ero1l", "Cyb5b", "Rac1", "Spta1", "Riok3", 
                   "Sox4", "Ran", "Dut", "Hmgb2", "Pim1", "Cd24a")

celltype_ery <- c("14/Stem"="HSC", "0"="Prog1",
                  "3"="Prog2", "10"="MEP", "8"="ProE",
                  "13"="Baso1", "11"="Baso2", "12/Ery"="Ery")

colors_ery <- c("HSC"="#826CC7", "Prog1"="#2B4285",
                "Prog2"="#8BBFE9", "MEP"="#AACA6A", "ProE"="#ECE592",
                "Baso1"="#F3A73B", "Baso2"="#EB3223", "Ery"="#7F170E") %>%
  { setNames(desaturate(lighten(.,0.2), 0.3), names(.)) }
```

## Functions

```{r methods}
boot_sparse <- function (M, num_boot=10) {
  ## the kronecker product captures all possible values we might fill in
  Y <- as(kronecker(M, Matrix(rep(1, num_boot), 1)), "dgCMatrix")
  
  ## let's provide column labels, using existing ones if present (default: "col1_1")
  colnames(Y) <- c(t(outer(colnames(M, do.NULL=FALSE), seq_len(num_boot), paste, sep='_')))
  
  ## generate multinomial samples and write them into the values
  Y@x <- as.double(unlist(sapply(colSums(M), function (n) {
    rmultinom(num_boot, n, rep(1,n))
  })))
  
  ## keep it sparsey sparse
  drop0(Y)
}

bootstrap_lui <- function (cts, M.groups, M.LU, M.gene, num_boot=100) {
  M.bs <- boot_sparse(M.groups, num_boot)
  LUI.bs <- ((M.LU %*% cts) %*% M.bs) / ((M.gene %*% cts) %*% M.bs)
  LUI.bs
}

bootstrap_wui <- function (cts, M.groups, M.WUI, M.gene, num_boot=100) {
  M.bs <- boot_sparse(M.groups, num_boot)
  WUI.bs <- ((M.WUI %*% cts) %*% M.bs) / ((M.gene %*% cts) %*% M.bs)
  WUI.bs
}

bootstrap_gene <- function (tpm, M.groups, M.gene, num_boot=100) {
  M.bs <- boot_sparse(M.groups, num_boot)
  D_scale <- colSums(M.bs) %>% { Diagonal(length(.), 1/., names=TRUE) }
  tpm.bs <- (M.gene %*% tpm) %*% M.bs %*% D_scale
  tpm.bs
}

plot_lui_bootstraps <- function (gene, df, keep_order=TRUE) {
  df %>%
    filter(gene == !!gene) %>%
    ggplot(aes(x=cell_type, y=lui, fill=cell_type, group=ct_b)) +
    geom_boxplot(outlier.shape=NA, size=0.2) +
    coord_flip() +
    scale_y_continuous(limits=c(0,1), breaks=seq(0,1,0.1)) +
    scale_fill_manual(values=colors_ery) +
    labs(x=NULL, y="LUI") +
    guides(fill=FALSE) +
    theme_bw()
}

plot_wui_bootstraps <- function (gene, df, keep_order=TRUE) {
  df %>%
    filter(gene == !!gene) %>%
    ggplot(aes(x=cell_type, y=wui, fill=cell_type, group=ct_b)) +
    geom_boxplot(outlier.shape=NA, size=0.2) +
    coord_flip() +
    scale_y_continuous(limits=c(0,1), breaks=seq(0,1,0.1)) +
    scale_fill_manual(values=colors_ery) +
    labs(x=NULL, y="WUI") +
    guides(fill=FALSE) +
    theme_bw()
}
```


# Data
## Loading
```{r load_data, message=FALSE}
sce <- readRDS("data/sce/hspcs.txs.full_annot.Rds")
```

## Preprocessing
```{r filter_data}
idx_cells <- sce$clusters %in% names(celltype_ery)
idx_txs <- rowData(sce) %>% { (!.$is_ipa) & (.$gene_name %in% genes_to_plot) }

D_cts_tpm <- assay(sce[,idx_cells], 'normcounts') %>%
  colSums() %>%
  { Diagonal(length(.), 1e6/.) }

sce %<>% `[`(idx_txs, idx_cells)

colData(sce)['cell_type'] <- factor(celltype_ery[sce$clusters], levels=celltype_ery)
```

## Design Matrices
```{r design_mats}
M_cell_celltype <- sce$cell_type %>% 
  fac2sparse %>% t

M_gene_tx <- rowData(sce)$gene_name %>% fac2sparse
M_lu_tx <- drop0(M_gene_tx %*% Diagonal(ncol(M_gene_tx), rowData(sce)$is_lu))

M_wu_tx <- rowData(sce) %>% as_tibble %>%
  group_by(gene_name) %>%
  mutate(utr_rank=rank(utr_position),
         utr_wt=(utr_rank - 1)/(max(utr_rank) - 1)) %>%
  ungroup() %>%
  pull(utr_wt, name=transcript_id) %>%
  { drop0(M_gene_tx %*% Diagonal(ncol(M_gene_tx), ., names=TRUE)) }
```

# Compute LUI Bootstraps
```{r compute_lui}
cts_tx_cell <- assay(sce, "normcounts")

ncells_gene_celltype <- ((M_gene_tx %*% cts_tx_cell) > 0) %*% M_cell_celltype

df_ncells <- ncells_gene_celltype %>%
  as.matrix %>% as_tibble(rownames="gene") %>%
  pivot_longer(cols=-1, names_to="cell_type", values_to="ncells") %>%
  mutate(cell_type=factor(cell_type, levels=celltype_ery))

lui_bs <- bootstrap_lui(cts_tx_cell, M_cell_celltype, 
                        M_lu_tx, M_gene_tx, num_boot=N_BOOTSTRAPS)

df_bs_lui <- lui_bs %>%
  as.matrix %>%
  as_tibble(rownames='gene') %>%
  pivot_longer(cols=-1, names_to="cell_type", names_pattern="^([^_]+)", values_to="lui") %>%
  mutate(cell_type=factor(cell_type, levels=celltype_ery)) %>%
  left_join(df_ncells, by=c("gene", "cell_type"))
```

# Compute WUI Bootstraps
```{r compute_wui}
wui_bs <- bootstrap_wui(cts_tx_cell, M_cell_celltype, 
                        M_wu_tx, M_gene_tx, num_boot=N_BOOTSTRAPS)

df_bs_wui <- wui_bs %>%
  as.matrix %>%
  as_tibble(rownames='gene') %>%
  pivot_longer(cols=-1, names_to="cell_type", names_pattern="^([^_]+)", values_to="wui") %>%
  mutate(cell_type=factor(cell_type, levels=celltype_ery)) %>%
  left_join(df_ncells, by=c("gene", "cell_type"))
```

# Compute Gene Bootstraps
```{r compute_tpms}
tpm_tx_cell <- cts_tx_cell %*% D_cts_tpm

tpm_bs <- bootstrap_gene(tpm_tx_cell, M_cell_celltype, 
                         M_gene_tx, num_boot=N_BOOTSTRAPS)

df_bs_tpm <- tpm_bs %>%
  as.matrix %>%
  as_tibble(rownames='gene') %>%
  pivot_longer(cols=-1, names_to="cell_type", names_pattern="^([^_]+)", values_to="tpm") %>%
  mutate(cell_type=factor(cell_type, levels=celltype_ery)) %>%
  left_join(df_ncells, by=c("gene", "cell_type"))
```

# Plots
## LUI
```{r plots_lui, fig.width=5, fig.height=3}
plot_lui_bootstraps_manual <- function (gene, df, keep_order=TRUE) {
  df_tmp <- df %>%
    filter(gene == !!gene) %T>%
    { axis_limits <<- c(min(.$lui), max(.$lui)) } %>%
    group_by(cell_type) %>%
    summarize(ci025=quantile(lui, 0.025), ci975=quantile(lui, 0.975),
              ci250=quantile(lui, 0.250), ci750=quantile(lui, 0.750),
              ci500=median(lui), ncells=dplyr::first(ncells), .groups='drop')
  
  
  ggplot(df_tmp, aes(x=cell_type, y=ci500, fill=cell_type)) +
    geom_errorbar(aes(ymin=ci025, ymax=ci975, color=cell_type), size=0.5,
                  position=position_dodge(width=0.95), width=0.5) +
    geom_crossbar(aes(ymin=ci250, ymax=ci750, color=cell_type), size=0.2,
                  position=position_dodge(width=0.95), width=0.8) +
    geom_crossbar(aes(ymin=ci500, ymax=ci500), size=0.1, color='black',
                  position=position_dodge(width=0.95), width=0.8) +
    geom_text(aes(label=ncells, y=axis_limits[1]), size=3,
              position=position_dodge(width=0.95), hjust=0.5, vjust=1.5) +
    #annotate(geom="text", x=0.65, y=axis_limits[1], label="N=", size=3,
    #         hjust=0.5, vjust=2) +
    #coord_flip(clip='off', ylim=axis_limits) +
    coord_cartesian(clip='off', ylim=axis_limits) +
    scale_x_discrete(position="bottom", guide=guide_axis(angle=90)) +
    scale_y_continuous(breaks=seq(0,1,0.1), expand=c(0.1,0,0.1,0)) +
    scale_fill_manual(values=colors_ery) +
    scale_color_manual(values=colors_ery) +
    labs(x=NULL, y="LUI", title=gene) +
    guides(fill=FALSE, color=FALSE) +
    theme_bw()
}

for (g in genes_to_plot) {
  print(g)
  plot_lui_bootstraps_manual(g, df_bs_lui) %>% print()
  
  ggsave(sprintf("output/figure4/fig4ab-bootstrap-lui-ery-celltypes-%s.pdf", g),
         width=5, height=3, dpi=300)
}
```

## WUI
```{r plots_wui, fig.width=5, fig.height=3}
plot_wui_bootstraps_manual <- function (gene, df, keep_order=TRUE) {
  df_tmp <- df %>%
    filter(gene == !!gene) %T>%
    { axis_limits <<- c(min(.$wui), max(.$wui)) } %>%
    group_by(cell_type) %>%
    summarize(ci025=quantile(wui, 0.025), ci975=quantile(wui, 0.975),
              ci250=quantile(wui, 0.250), ci750=quantile(wui, 0.750),
              ci500=median(wui), ncells=dplyr::first(ncells), .groups='drop')
  
  if (axis_limits[[2]] - axis_limits[[1]] < MIN_UI_RANGE) {
    mid_axis <- mean(axis_limits)
    if (mid_axis < MIN_UI_RANGE/2) {
      axis_limits <- c(0, MIN_UI_RANGE)
    } else if (mid_axis > 1 - MIN_UI_RANGE/2) {
      axis_limits <- c(1-MIN_UI_RANGE, 1)
    } else {
      axis_limits <- c(mid_axis - MIN_UI_RANGE/2, mid_axis + MIN_UI_RANGE/2)
    }
  }
  
  ggplot(df_tmp, aes(x=cell_type, y=ci500, fill=cell_type)) +
    geom_errorbar(aes(ymin=ci025, ymax=ci975, color=cell_type), size=0.5,
                  position=position_dodge(width=0.95), width=0.5) +
    geom_crossbar(aes(ymin=ci250, ymax=ci750, color=cell_type), size=0.2,
                  position=position_dodge(width=0.95), width=0.8) +
    geom_crossbar(aes(ymin=ci500, ymax=ci500), size=0.1, color='black',
                  position=position_dodge(width=0.95), width=0.8) +
    geom_text(aes(label=ncells, y=axis_limits[1]), size=3,
              position=position_dodge(width=0.95), hjust=0.5, vjust=1.5) +
    #annotate(geom="text", x=0.65, y=axis_limits[1], label="N=", size=3,
    #         hjust=0.5, vjust=2) +
    #coord_flip(clip='off', ylim=axis_limits) +
    coord_cartesian(clip='off', ylim=axis_limits) +
    scale_x_discrete(position="bottom", guide=guide_axis(angle=90)) +
    scale_y_continuous(breaks=seq(0,1,0.1), expand=c(0.1,0,0.1,0)) +
    scale_fill_manual(values=colors_ery) +
    scale_color_manual(values=colors_ery) +
    labs(x=NULL, y="WUI", title=gene) +
    guides(fill=FALSE, color=FALSE) +
    theme_bw()
}

for (g in genes_to_plot) {
  print(g)
  plot_wui_bootstraps_manual(g, df_bs_wui) %>% print()
  
  ggsave(sprintf("output/figure4/fig4ab-bootstrap-wui-ery-celltypes-%s.pdf", g),
         width=5, height=3, dpi=300)
}
```

## Gene TPM
```{r plots_tpms, fig.width=5, fig.height=3}
plot_tpm_bootstraps_manual <- function (gene, df, keep_order=TRUE) {
  df_tmp <- df %>%
    filter(gene == !!gene) %T>%
    #{ axis_limits <<- c(min(.$tpm), max(.$tpm)) } %>%
    { axis_limits <<- c(0, max(.$tpm)) } %>%
    group_by(cell_type) %>%
    summarize(ci025=quantile(tpm, 0.025), ci975=quantile(tpm, 0.975),
              ci250=quantile(tpm, 0.250), ci750=quantile(tpm, 0.750),
              ci500=median(tpm), ncells=dplyr::first(ncells), .groups='drop')
  
  
  ggplot(df_tmp, aes(x=cell_type, y=ci500, fill=cell_type)) +
    geom_errorbar(aes(ymin=ci025, ymax=ci975, color=cell_type), size=0.5,
                  position=position_dodge(width=0.95), width=0.5) +
    geom_crossbar(aes(ymin=ci250, ymax=ci750, color=cell_type), size=0.2,
                  position=position_dodge(width=0.95), width=0.8) +
    geom_crossbar(aes(ymin=ci500, ymax=ci500), size=0.1, color='black',
                  position=position_dodge(width=0.95), width=0.8) +
    geom_text(aes(label=ncells, y=axis_limits[1]), size=3,
              position=position_dodge(width=0.95), hjust=0.5, vjust=1.5) +
    #annotate(geom="text", x=0.65, y=axis_limits[1], label="N=", size=3,
    #         hjust=0.5, vjust=2) +
    #coord_flip(clip='off', ylim=axis_limits) +
    coord_cartesian(clip='off', ylim=axis_limits) +
    scale_x_discrete(position="bottom", guide=guide_axis(angle=90)) +
    scale_y_continuous(expand=c(0.1,0,0.1,0)) +
    scale_fill_manual(values=colors_ery) +
    scale_color_manual(values=colors_ery) +
    labs(x=NULL, y="TPM", title=gene) +
    guides(fill=FALSE, color=FALSE) +
    theme_bw()
}

for (g in genes_to_plot) {
  print(g)
  plot_tpm_bootstraps_manual(g, df_bs_tpm) %>% print()
  
  ggsave(sprintf("output/figure4/fig4ab-bootstrap-tpm-ery-celltypes-%s.pdf", g),
         width=5, height=3, dpi=300)
}
```

---

# Session Info

```{r sesh_info, echo=FALSE}
sessionInfo()
```
