---
title: "Cluster Plots"
author: "Mervin Fansler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    toc: true
    toc_depth: 2
---

# Purpose 

Here we compare clustering results using CellRanger versus scUTRquant derived counts
from the 10X Chromium 3'-end demonstration data (v2 and v3). This gauges consistency 
between gene-level count pipelines, with a focus on potential downstream differences.

# Libraries
```{r libs, message=FALSE, warning=FALSE}
library(SingleCellExperiment)
library(bluster)
library(tidyverse)
library(magrittr)
library(cowplot)
library(ggalluvial)
library(pheatmap)
library(ggbeeswarm)
```

## Parameters
```{r set_params}
set.seed(20210927)
```

## Functions

```{r methods}
cluster_louvain <- function (sce) {
  clusterRows(reducedDim(sce, "PCA"), NNGraphParam(cluster.fun="louvain"))
}

plot_cluster_areas <- function (tbl, label_x="CellRanger", label_y="scUTRquant",
                                show_values=FALSE,
                                plotCols=c('#FFFFFF', colorRampPalette(RColorBrewer::brewer.pal(9, "Blues")[2:7])(10))) {
  tibble(x=rep(cumsum(colSums(tbl)) - 0.5*colSums(tbl), each=nrow(tbl)),
         y=rep(cumsum(rowSums(tbl)) - 0.5*rowSums(tbl), times=ncol(tbl)),
         height=rep(rowSums(tbl), times=ncol(tbl)),
         width=rep(colSums(tbl), each=nrow(tbl)),
         values=as.vector(tbl/rowSums(tbl))) %>%
    ggplot(aes(x=x, y=y)) +
    geom_tile(aes(width=width, height=height, fill=values), color='black') +
    { if (show_values) geom_text(aes(label=round(values, 2))) } +
    scale_fill_gradientn(colors=plotCols) +
    labs(x=label_x, y=label_y) +
    coord_equal() +
    theme_minimal_grid()
}

make_cluster_df <- function (cr, sq, idx_cr=NULL, idx_sq=NULL) {
  if (!is.null(idx_cr)) { 
    cr <- fct_recode(cr, !!!setNames(as.character(idx_cr),levels(cr)))
  }
  if (!is.null(idx_sq)) {
    sq <- fct_recode(sq, !!!setNames(as.character(idx_sq),levels(sq)))
  }
  tibble(cell=seq_along(cr), 
         CellRanger=factor(cr, levels=sort(as.numeric(unique(cr)))),
         scUTRquant=factor(sq, levels=sort(as.numeric(unique(sq))))) %>%
    pivot_longer(-1, names_to="method", values_to="cluster")
}
```

# Data
## Loading
```{r load_data, message=FALSE}
df_counts <- readRDS("data/sce/processed/df_counts.all_genes.pca.Rds")

## validate cell id orders
df_counts %>%
  transmute(sample_id, matched=map2_lgl(sce_10x, sce_sq, ~ all(colnames(.x)==colnames(.y)))) %>%
  deframe() %>%
  stopifnot()
```

## Preprocessing
```{r prepare_data}
df_clusters <- df_counts %>%
  
  ## cluster
  mutate(clusters_sq=map(sce_sq, cluster_louvain),
         clusters_10x=map(sce_10x, cluster_louvain)) %>%
  
  transmute(sample_id, clusters_10x, clusters_sq,
            n_clusters_10x=map_int(clusters_10x, nlevels),
            n_clusters_sq=map_int(clusters_sq, nlevels),
            ari=map2_dbl(clusters_sq, clusters_10x, aricode::ARI),
            nmi=map2_dbl(clusters_sq, clusters_10x, aricode::NMI),
            nvi=map2_dbl(clusters_sq, clusters_10x, aricode::NVI),
            table=map2(clusters_10x, clusters_sq, table))
```

# Analysis

## Cluster Counts
```{r}
df_clusters %>%
  select(sample_id, n_clusters_10x, n_clusters_sq, ari) %>%
  knitr::kable()
```

### ARI Bar
```{r ari_bar, fig.width=2, fig.height=4}
n_lab <- sprintf("N = %d", nrow(df_clusters))

df_clusters %>%
  ggplot(aes(x=n_lab, y=ari)) + 
  stat_summary(fun='mean', geom='bar', width=0.4, fill='lightgrey', color='black') +
  stat_summary(fun.data=mean_se, geom='errorbar', width=0.3) +
  geom_quasirandom(width=0.1, size=1) +
  scale_y_continuous(limits=c(0,1), expand=expansion(add=c(0,0))) +
  labs(x=NULL, y="Adjusted Rand Index") +
  theme_bw()

ggsave("output/figure1/sup1D-clusters-ari.pdf", 
       width=2, height=4, dpi=300)
```

## Raw Cluster Plots
```{r raw_clusters, fig.width=5, fig.height=4}
plotCols <- colorRampPalette(RColorBrewer::brewer.pal(9, "Blues")[1:7])(10)
for (i in seq(nrow(df_clusters))) {
  tbl <- df_clusters$table[[i]]
  idx_row <- order(rowSums(tbl), decreasing=TRUE)
  idx_col <- order(colSums(tbl), decreasing=TRUE)
  (tbl/rowSums(tbl)) %>%
    `[`(idx_row, idx_col) %>%
  pheatmap(display_numbers=FALSE, color=plotCols, main=df_clusters$sample_id[[i]],
           cluster_rows=FALSE, cluster_cols=FALSE, angle_col=0)
}
```

## Raw Cluster Areas
```{r raw_areas, fig.width=5, fig.height=4}
for (i in seq(nrow(df_clusters))) {
  tbl <- df_clusters$table[[i]]
  idx_row <- order(rowSums(tbl), decreasing=TRUE)
  idx_col <- order(colSums(tbl), decreasing=TRUE)
  tbl[idx_row, idx_col] %>%
    plot_cluster_areas %>%
    print
}
```

```{r ordered_1, fig.width=5, fig.height=4}
idx_row <- c(7,2,4,3,5,6,1,8,9)
idx_col <- c(7,2,4,3,5,6,1,8,9)
tbl <- df_clusters$table[[1]]
t(tbl[idx_row, idx_col]) %>%
  plot_cluster_areas(show_values=FALSE) %>%
  { . + labs(title=paste(df_clusters$sample_id[[1]], "Clusters")) } %>%
  print
```


```{r plot_alluvial_default_1, fig.width=4, fig.height=4}
df_clusters[1,] %>%
  { make_cluster_df(cr=.$clusters_10x[[1]], sq=.$clusters_sq[[1]],
                    idx_cr=c(7,2,4,3,5,6,1,8,9),
                    idx_sq=c(7,2,4,3,5,6,1,8,9)) } %>%

  ggplot(aes(x=method, stratum=cluster, alluvium=cell, label=cluster)) +
  geom_flow(aes(fill=cluster), width=0.2, ) +
  geom_stratum(fill='lightgrey', alpha=0.5, width=0.2) +
  geom_text(stat='stratum') +
  scale_x_discrete(limits = c("CellRanger", "scUTRquant"), expand = c(0.1, 0.1)) +
  scale_fill_manual(values=unname(palette.colors(9))) +
  labs(x="Method", y="Cells") +
  guides(fill=FALSE) +
  theme_cowplot()
```


```{r plot_alluvial_brbg, fig.width=5, fig.height=5}
set.seed(3)
colors <- sample(RColorBrewer::brewer.pal(11, 'BrBG')[c(1:4,7:11)])
df_clusters[1,] %>%
  { make_cluster_df(cr=.$clusters_10x[[1]], sq=.$clusters_sq[[1]],
                    idx_cr=c(7,2,4,3,5,6,1,8,9),
                    idx_sq=c(7,2,4,3,5,6,1,8,9)) } %>%

  ggplot(aes(x=method, stratum=cluster, alluvium=cell, label=cluster)) +
  geom_flow(aes(fill=cluster), width=0.2) +
  geom_stratum(fill='lightgrey', alpha=0.2, width=0.2) +
  geom_text(stat='stratum') +
  scale_x_discrete(limits = c("CellRanger", "scUTRquant"), expand = c(0.1, 0.1)) +
  scale_y_continuous(expand=c(0,0)) +
  scale_fill_manual(values=colors) +
  labs(x="Method", y="Cells") +
  guides(fill=FALSE) +
  theme_cowplot()

ggsave("output/figure1/fig1D-clusters-alluvial-heart_1k_v2_fastq-BrBG.pdf", 
       width=5, height=5, dpi=300)
```


```{r plot_alluvial_spectral, fig.width=5, fig.height=5}
df_clusters[1,] %>%
  { make_cluster_df(cr=.$clusters_10x[[1]], sq=.$clusters_sq[[1]],
                    idx_cr=c(7,2,4,3,5,6,1,8,9),
                    idx_sq=c(7,2,4,3,5,6,1,8,9)) } %>%

  ggplot(aes(x=method, stratum=cluster, alluvium=cell, label=cluster)) +
  geom_flow(aes(fill=cluster), alpha=0.45, width=0.2) +
  geom_stratum(fill='lightgrey', alpha=0.4, width=0.2) +
  geom_text(stat='stratum') +
  scale_x_discrete(limits = c("CellRanger", "scUTRquant"), expand = c(0.1, 0.1)) +
  scale_y_continuous(expand=c(0,0)) +
  scale_fill_manual(values=RColorBrewer::brewer.pal(10, 'Spectral')[c(1:5,7:10)]) +
  labs(x="Method", y="Cells") +
  guides(fill=FALSE) +
  theme_cowplot()

ggsave("output/figure1/fig1D-clusters-alluvial-heart_1k_v2_fastq-spectral.pdf", 
       width=5, height=5, dpi=300)
```

## heart_1k_v3
```{r ordered_2, fig.width=5, fig.height=4}
idx_row <- c(1,2,3,4,5,11,6,7,9,8,10,12,13,14)
idx_col <- c(1,2,3,4,5,6,7,9,8,10,11,12,13)
tbl <- df_clusters$table[[2]]
t(tbl[idx_row, idx_col]) %>%
  plot_cluster_areas(show_values=FALSE) %>%
  { . + labs(title=paste(df_clusters$sample_id[[2]], "Clusters")) } %>%
  print
  
```

```{r plot_alluvial_default_2, fig.width=5, fig.height=5}
df_clusters[2,] %>%
  { make_cluster_df(cr=.$clusters_10x[[1]], sq=.$clusters_sq[[1]],
                    idx_cr=c(1,2,3,4,5,11,6,7,9,8,10,12,13,14),
                    idx_sq=c(1,2,3,4,5,6,7,9,8,10,11,12,13)) } %>%

  ggplot(aes(x=method, stratum=cluster, alluvium=cell, label=cluster)) +
  geom_flow(aes(fill=cluster), width=0.2, ) +
  geom_stratum(fill='lightgrey', alpha=0.5, width=0.2) +
  geom_text(stat='stratum') +
  scale_x_discrete(limits = c("CellRanger", "scUTRquant"), expand = c(0.1, 0.1)) +
  scale_fill_manual(values=colorRampPalette(palette.colors(9))(14)) +
  labs(x="Method", y="Cells") +
  guides(fill=FALSE) +
  theme_cowplot()

ggsave("output/figure1/fig1D-clusters-alluvial-heart_1k_v3_fastq-default.pdf", 
       width=5, height=5, dpi=300)
```


```{r plot_alluvial_spectral_2, fig.width=5, fig.height=5}
colors <- colorRampPalette(RColorBrewer::brewer.pal(10, 'Spectral')[c(1:5,7:10)])(14)
df_clusters[2,] %>%
  { make_cluster_df(cr=.$clusters_10x[[1]], sq=.$clusters_sq[[1]],
                    idx_cr=c(1,2,3,4,5,11,6,7,9,8,10,12,13,14),
                    idx_sq=c(1,2,3,4,5,6,7,9,8,10,11,12,13)) } %>%

  ggplot(aes(x=method, stratum=cluster, alluvium=cell, label=cluster)) +
  geom_flow(aes(fill=cluster), width=0.2, ) +
  geom_stratum(fill='lightgrey', alpha=0.5, width=0.2) +
  geom_text(stat='stratum') +
  scale_x_discrete(limits = c("CellRanger", "scUTRquant"), expand = c(0.1, 0.1)) +
  scale_fill_manual(values=colors) +
  labs(x="Method", y="Cells") +
  guides(fill=FALSE) +
  theme_cowplot()

ggsave("output/figure1/fig1D-clusters-alluvial-heart_1k_v3_fastq-spectral.pdf", 
       width=5, height=5, dpi=300)
```


---

# Runtime Details
## Session Info
```{r sesh_info, echo=FALSE}
sessionInfo()
```

## Conda Environment
```{bash comment="", echo=FALSE}
if ! command -v conda &> /dev/null
then
  echo "Conda not detected."
elif [ -z "${CONDA_PREFIX}" ]
then
  echo "No active Conda environment."
else
  echo "## Conda Environment YAML"
  conda env export
fi
```
