---
title: "Compare CellRanger vs scUTRquant DimReds"
author: "Mervin Fansler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    code_folding: show
    toc: true
    toc_float: true
---

# Purpose

<!--
Explain why this document exists. 
What are you trying to figure out?
What are your expectations prior to the analysis?
-->
Here we plot PCA and UMAP dimensionality reductions of the UMI count data from 
CellRanger and scUTRquant for comparison. We expect highly similar projections 
since the underlying data is identical.

# Initialization

## Libraries
```{r libs, message=FALSE, warning=FALSE}
library(magrittr)
library(tidyverse)
library(cowplot)
library(Matrix)
library(HDF5Array)
library(SingleCellExperiment)
library(scater)
```

## Parameters
```{r set_params}
set.seed(20210818)
BiocParallel::register(BiocParallel::MulticoreParam(12))
```

## Functions

```{r methods}
plot_dimred_compare <- function (sce_x, sce_y, dimred="PCA",
                                 label_x="CellRanger", label_y="scUTRquant") {
  gx <- plotReducedDim(sce_x, dimred)
  gy <- plotReducedDim(sce_y, dimred)
  glabx <- ggdraw() + 
    draw_label(label_x, fontface='bold', size=16, hjust=0.36, vjust=0.5) +
    theme(plot.margin=margin(0, 0, 6, 0))
  glaby <- ggdraw() + 
    draw_label(label_y, fontface='bold', size=16, hjust=0.36, vjust=0.5) +
    theme(plot.margin=margin(0, 0, 6, 0))
  plot_grid(gx, gy, glabx, glaby, nrow=2, rel_heights=c(1, 0.05))
}
```

# Data
## Loading
```{r load_data, message=FALSE}
df_counts_all <- readRDS("data/sce/processed/df_counts.all_genes.pca.Rds")
```

## Preprocessing
### Compute UMAP Coordinates
```{r prepare_data, message=FALSE}
df_counts_all %<>%
  mutate(sce_sq=map(sce_sq, runUMAP),
         sce_10x=map(sce_10x, runUMAP))
```

# Analysis
## PCA

```{r plot_pca, fig.width=8, fig.height=4}
df_counts_all %>%
  transmute(sample_id, g=map2(sce_10x, sce_sq, plot_dimred_compare, dimred="PCA")) %>%
  deframe() %>% {
    for (id in names(.)) {
      title <- ggdraw() + 
        draw_label(id, fontface='bold', size=20, x=0.5, hjust=0.5) +
        theme(plot.margin=margin(0, 0, 0, 0))
      g <- plot_grid(title, .[[id]], ncol=1, rel_heights=c(0.1,1))
      
      ## print locally in this document
      print(g)
      
      ## save to PDFs
      # ggsave(sprintf("output/supplement/sup2-pca-compare-%s.pdf", id), 
      #        .[[id]], width=5, height=4, dpi=300)
    }
  }
```

## UMAP

```{r plot_umap, fig.width=8, fig.height=4}
df_counts_all %>%
  transmute(sample_id, g=map2(sce_10x, sce_sq, plot_dimred_compare, dimred="UMAP")) %>%
  deframe() %>% {
    for (id in names(.)) {
      title <- ggdraw() + 
        draw_label(id, fontface='bold', size=20, x=0.5, hjust=0.5) +
        theme(plot.margin=margin(0, 0, 0, 0))
      g <- plot_grid(title, .[[id]], ncol=1, rel_heights=c(0.1,1))
      
      ## print locally in this document
      print(g)
      
      ## save to PDFs
      # ggsave(sprintf("output/supplement/sup2-pca-compare-%s.pdf", id), 
      #        .[[id]], width=5, height=4, dpi=300)
    }
  }
```

# Conclusion

There is some structural similarity in the projections, but it is not immediately 
evident that exactly the same clusters are present. It is also not clear whether
one mode is any more articulated than the other. In all samples, the first two 
PCs for CellRanger capture more variance than those for scUTRquant.

---

# Runtime Details
## Session Info
```{r sesh_info, echo=FALSE}
sessionInfo()
```

## Conda Environment
```{bash conda, comment="", echo=FALSE}
if ! command -v conda &> /dev/null
then
  echo "Conda not detected."
elif [ -z "${CONDA_PREFIX}" ]
then
  echo "No active Conda environment."
else
  echo "## Conda Environment YAML"
  conda env export
fi
```
