---
title: "Compare CellRanger vs scUTRquant HVGs"
author: "Mervin Fansler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    code_folding: show
    toc: true
    toc_float: true
---

# Purpose

<!--
Explain why this document exists. 
What are you trying to figure out?
What are your expectations prior to the analysis?
-->
This document compares the sets of highest-variance genes across CellRanger and 
scUTRquant gene UMI counts. Currently, the clusters are not in perfect agreement
and so this is being examined in order to determine if the sets of genes being selected
are playing a significant role.

# Initialization

## Libraries
```{r libs, message=FALSE, warning=FALSE}
library(magrittr)
library(tidyverse)
library(cowplot)
library(Matrix)
library(HDF5Array)
library(SingleCellExperiment)
library(scran)
```

## Parameters
```{r set_params}
set.seed(20210818)
BiocParallel::register(BiocParallel::MulticoreParam(12))
```

## Functions

```{r methods}
get_pc_df <- function (sce_x, sce_y, pc="PC1",
                        label_x="CellRanger", label_y="scUTRquant") {
  df_rot_x <- reducedDim(sce_x, "PCA") %>% 
    attr("rotation") %>% `[`(,pc) %>% 
    enframe("gene", label_x)
  df_rot_y <- reducedDim(sce_y, "PCA") %>% 
    attr("rotation") %>% `[`(,pc) %>% 
    enframe("gene", label_y)
  full_join(df_rot_x, df_rot_y, by="gene") %>% 
    mutate(across(where(is.numeric), coalesce, 0))
}

plot_weights_compare <- function (df) {
  max_abs_val <- max(abs(as.matrix(df[,-1])))
  ggplot(df, aes(x=CellRanger, y=scUTRquant)) +
    geom_abline(slope=c(1,-1), 
                intercept=c(0,0), linetype='dashed') +
    geom_abline(slope=c(1,1,-1,-1), 
                intercept=c(-0.05, 0.05, -0.05, 0.05), 
                linetype='dashed', color='lightgrey') +
    geom_point(size=1, alpha=0.8, pch=16) +
    coord_cartesian(xlim=c(-max_abs_val, max_abs_val),
                    ylim=c(-max_abs_val, max_abs_val)) +
    theme_bw()
}
```

# Data
## Loading
```{r load_data, message=FALSE}
df_counts_all <- readRDS("data/sce/processed/df_counts.all_genes.pca.Rds")
```

## Preprocessing
### Extract PC Dataframes
```{r prepare_data, message=FALSE}
df_pcs <- df_counts_all %>%
  transmute(sample_id, 
            PC1=map2(sce_10x, sce_sq, get_pc_df, pc="PC1"),
            PC2=map2(sce_10x, sce_sq, get_pc_df, pc="PC2"))
```

# Analysis
## Plot PC1 Weights

```{r plot_pc1, fig.width=4, fig.height=4}
df_pcs %>%
  transmute(sample_id, g=map(PC1, plot_weights_compare)) %>%
  deframe() %>% {
    for (id in names(.)) {
      ## print locally in this document
      print(.[[id]] + ggtitle(id))
      
      ## save to PDFs
      # ggsave(sprintf("output/supplement/sup2-pca-compare-%s.pdf", id), 
      #        .[[id]], width=5, height=4, dpi=300)
    }
  }
```

### List Genes
```{r list_p1_genes}
df_pcs %>%
  transmute(sample_id, g=map(PC1, function (df) {
    df %>%
      #filter(CellRanger == 0 | scUTRquant == 0 | abs(CellRanger - scUTRquant) > 0.05) %>%
      filter(abs(CellRanger - scUTRquant) > 0.1) %>%
      pull(gene)
  }))
```

## Plot PC2 Weights

```{r plot_pc2, fig.width=4, fig.height=4}
df_pcs %>%
  transmute(sample_id, g=map(PC2, plot_weights_compare)) %>%
  deframe() %>% {
    for (id in names(.)) {
      ## print locally in this document
      print(.[[id]] + ggtitle(id))
      
      ## save to PDFs
      # ggsave(sprintf("output/supplement/sup2-pca-compare-%s.pdf", id), 
      #        .[[id]], width=5, height=4, dpi=300)
    }
  }
```

# Conclusion

There are major discrepancies in the PC weightings, especially in the neuronal data.
We need to dig deeper into which genes have very large differences.

---

# Runtime Details
## Session Info
```{r sesh_info, echo=FALSE}
sessionInfo()
```

## Conda Environment
```{bash conda, comment="", echo=FALSE}
if ! command -v conda &> /dev/null
then
  echo "Conda not detected."
elif [ -z "${CONDA_PREFIX}" ]
then
  echo "No active Conda environment."
else
  echo "## Conda Environment YAML"
  conda env export
fi
```
