---
title: "Compare CellRanger vs scUTRquant Clustering"
author: "Mervin Fansler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    toc: true
    toc_depth: 2
---

# Purpose 

Here we compare clustering results using CellRanger versus scUTRquant derived counts
from the 10X Chromium 3'-end demonstration data (v2 and v3). This gauges consistency 
between gene-level count pipelines, with a focus on potential downstream differences.

# Libraries
```{r libs, message=FALSE, warning=FALSE}
library(SingleCellExperiment)
library(DropletUtils)
library(scran)
library(scater)
library(bluster)
library(tidyverse)
library(magrittr)
library(cowplot)
library(Matrix)
```

# Load and Preprocess Data
```{r load_data, message=FALSE, warning=FALSE}
set.seed(20210723)
SAMPLE_SHEET="metadata/counts_sample_sheet.csv"

## conform 10X cell ids to match scUTRquant cell_id
conform_cell_ids <- function (sample_id, sce) {
  colData(sce) %<>%
    as_tibble %>%
    mutate(bx=str_extract(Barcode, "^[ACGT]{16}"),
           cell_id=str_c(sample_id, bx, sep='_')) %>%
    select(cell_id, bx) %>%
    set_rownames(.$cell_id) %>%
    DataFrame()
  sce
}

## summarize scUTRquant transcript counts to gene counts
txs_to_genes <- function (sce) {
  M_genes_txs <- rowData(sce)$gene_id %>%
    fac2sparse %>%
    set_rownames(str_extract(rownames(.), "^ENSMUSG[0-9]+"))
  SingleCellExperiment(assays=list(counts=M_genes_txs %*% counts(sce)),
                       colData=colData(sce))
}

intersect_cell_ids <- function (sce1, sce2) { 
  intersect(colnames(sce1), colnames(sce2))
}

## computes the diversity of genes represented per cell
## 2^(entropy)
compute_effective_gene_count <- function (sce) {
  cts <- counts(sce) %>% Matrix(sparse=TRUE)
  pct <- (cts %*% Diagonal(ncol(cts), 1/colSums(cts)))
  lpct <- pct
  lpct@x <- log2(lpct@x)
  2^(-colSums(pct*lpct))
}

filter_cells <- function (sce, idx) { sce[,idx] }

## exclude cells with low diversity
## this often includes both high MT and high Hb(a|b) genes
filter_low_diversity_cells <- function (sce, min_eff_genes=100) {
  n_eff_genes <- compute_effective_gene_count(sce)
  sce[, n_eff_genes >= min_eff_genes]
}

## adds PCA to SCE object
add_pca <- function (sce) {
  tmp_clusters <- quickCluster(sce)
  sce <- computeSumFactors(sce, clusters=tmp_clusters)
  sce <- logNormCounts(sce)
  dec <- modelGeneVar(sce)
  hvgs <- getTopHVGs(dec, n=1000)
  runPCA(sce, ncomponents=20, subset_row=hvgs)
}

df_counts <- read_csv(SAMPLE_SHEET) %>%
  ## read SCE files
  mutate(sce_sq=map(file_scutrquant, readRDS),
         sce_10x=map(file_cellranger, read10xCounts)) %>%
  
  ## adjust cell_ids
  mutate(sce_10x=map2(sample_id, sce_10x, conform_cell_ids)) %>%
  
  ## summarize to gene counts
  mutate(sce_sq=map(sce_sq, txs_to_genes)) %>%
  
  ## filter low-diversity cells
  mutate(sce_sq=map(sce_sq, filter_low_diversity_cells),
         sce_10x=map(sce_10x, filter_low_diversity_cells)) %>%
  
  ## filter to common cells
  mutate(idx_cells=map2(sce_sq, sce_10x, intersect_cell_ids)) %>%
  mutate(sce_sq=map2(sce_sq, idx_cells, filter_cells),
         sce_10x=map2(sce_10x, idx_cells, filter_cells)) %>%

  ## add PCA to SCEs
  mutate(sce_sq=map(sce_sq, add_pca),
         sce_10x=map(sce_10x, add_pca))
```

# All Genes
### Export
```{r export_basic} 
saveRDS(df_counts, "data/sce/processed/df_counts.all_genes.pca.Rds")
```

## Clustering
### Louvain
```{r compute_louvain_clusters}
cluster_louvain <- function (sce) {
  clusterRows(reducedDim(sce, "PCA"), NNGraphParam(cluster.fun="louvain"))
}

df_counts %>%
  ## cluster
  mutate(clusters_sq=map(sce_sq, cluster_louvain),
         clusters_10x=map(sce_10x, cluster_louvain)) %>%
  
  ## compute adjusted Rand index
  transmute(sample_id, n_cells=map_int(idx_cells, length),  
            n_clusters_10x=map_int(clusters_10x, nlevels),
            n_clusters_sq=map_int(clusters_sq, nlevels),
            ari=map2_dbl(clusters_sq, clusters_10x, aricode::ARI),
            nmi=map2_dbl(clusters_sq, clusters_10x, aricode::NMI),
            nvi=map2_dbl(clusters_sq, clusters_10x, aricode::NVI)) %>%
  
  knitr::kable()
```

### Walktrap
```{r compute_clusters}
cluster_walktrap<- function (sce) {
  clusterRows(reducedDim(sce, "PCA"), NNGraphParam(cluster.fun="walktrap"))
}

df_counts %>%
  ## cluster
  mutate(clusters_sq=map(sce_sq, cluster_walktrap),
         clusters_10x=map(sce_10x, cluster_walktrap)) %>%
  
  ## compute adjusted Rand index
  transmute(sample_id, n_cells=map_int(idx_cells, length),  
            n_clusters_10x=map_int(clusters_10x, nlevels),
            n_clusters_sq=map_int(clusters_sq, nlevels),
            ari=map2_dbl(clusters_sq, clusters_10x, aricode::ARI),
            nmi=map2_dbl(clusters_sq, clusters_10x, aricode::NMI),
            nvi=map2_dbl(clusters_sq, clusters_10x, aricode::NVI)) %>%
  
  knitr::kable()
```

# Common Genes
```{r common_genes}
intersect_gene_ids <- function (sce1, sce2) { 
  intersect(rownames(sce1), rownames(sce2))
}

filter_features <- function (sce, idx) { sce[idx,] }

df_counts <- df_counts %>%
  ## filter to common cells
  mutate(idx_genes=map2(sce_sq, sce_10x, intersect_gene_ids)) %>%
  mutate(sce_sq=map2(sce_sq, idx_genes, filter_features),
         sce_10x=map2(sce_10x, idx_genes, filter_features)) %>%

  ## add PCA to SCEs
  mutate(sce_sq=map(sce_sq, add_pca),
         sce_10x=map(sce_10x, add_pca))
```

### Export
```{r export_common} 
saveRDS(df_counts, "data/sce/processed/df_counts.common_genes.pca.Rds")
```

## Clustering
### Louvain
```{r compute_louvain_clusters_common}
df_counts %>%
  ## cluster
  mutate(clusters_sq=map(sce_sq, cluster_louvain),
         clusters_10x=map(sce_10x, cluster_louvain)) %>%
  
  ## compute adjusted Rand index
  transmute(sample_id, n_cells=map_int(idx_cells, length),  
            n_clusters_10x=map_int(clusters_10x, nlevels),
            n_clusters_sq=map_int(clusters_sq, nlevels),
            ari=map2_dbl(clusters_sq, clusters_10x, aricode::ARI),
            nmi=map2_dbl(clusters_sq, clusters_10x, aricode::NMI),
            nvi=map2_dbl(clusters_sq, clusters_10x, aricode::NVI)) %>%
  
  knitr::kable()
```

### Walktrap
```{r compute_clusters_common}
df_counts %>%
  ## cluster
  mutate(clusters_sq=map(sce_sq, cluster_walktrap),
         clusters_10x=map(sce_10x, cluster_walktrap)) %>%
  
  ## compute adjusted Rand index
  transmute(sample_id, n_cells=map_int(idx_cells, length),  
            n_clusters_10x=map_int(clusters_10x, nlevels),
            n_clusters_sq=map_int(clusters_sq, nlevels),
            ari=map2_dbl(clusters_sq, clusters_10x, aricode::ARI),
            nmi=map2_dbl(clusters_sq, clusters_10x, aricode::NMI),
            nvi=map2_dbl(clusters_sq, clusters_10x, aricode::NVI)) %>%
  
  knitr::kable()
```


---

# Session Info
```{r sesh_info, echo=FALSE}
sessionInfo()
```
