---
title: "Compare CellRanger vs scUTRquant Clustering - Multimapping"
author: "Mervin Fansler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    toc: true
    toc_depth: 2
---

# Purpose 

Here we compare clustering results using CellRanger versus scUTRquant derived counts
from the 10X Chromium 3'-end demonstration data (v2 and v3). This is a diagnostic 
run, not intended for publication or downstream analysis. We aim to understand what
genes are most driving the differences observed in clustering.

# Setup
## Libraries
```{r libs, message=FALSE, warning=FALSE}
library(SingleCellExperiment)
library(DropletUtils)
library(scran)
library(scater)
library(bluster)
library(tidyverse)
library(magrittr)
library(cowplot)
library(Matrix)
```

## Parameters
```{r params}
set.seed(20210723)
BiocParallel::register(BiocParallel::MulticoreParam(12))
SAMPLE_SHEET="metadata/counts_multimap_sample_sheet.csv"

N_HVGS=1000 # controls genes used for PCA
N_PCS=20 # number of principal components to compute
MIN_EFF_GENES=100 # requires cells to be diverse in gene
MAX_RATIO=Inf # controls maximum ratio difference between gene counts
```


## Functions
```{r functions, message=FALSE, warning=FALSE}
## load multimapping table
read_multimap_tsv <- function (file) {
  read_tsv(file, skip=1, col_types='c__l_',
           col_names=c("gene_symbol", "is_multimapping")) %>%
    group_by(gene_symbol) %>%
    summarize(has_multimapping=any(is_multimapping), .groups='drop')
}

## conform 10X cell ids to match scUTRquant cell_id
conform_cell_ids <- function (sample_id, sce) {
  colData(sce) %<>%
    as_tibble %>%
    mutate(bx=str_extract(Barcode, "^[ACGT]{16}"),
           cell_id=str_c(sample_id, bx, sep='_')) %>%
    select(cell_id, bx) %>%
    set_rownames(.$cell_id) %>%
    DataFrame()
  sce
}

## conform to unversioned ENSEMBL gene_id
conform_gene_ids <- function (sce) {
  set_rownames(sce, str_extract(rownames(sce), "^ENSMUSG[0-9]+"))
}

intersect_cell_ids <- function (sce1, sce2) { 
  intersect(colnames(sce1), colnames(sce2))
}

## computes the diversity of genes represented per cell
## 2^(entropy)
compute_effective_gene_count <- function (sce) {
  cts <- counts(sce) %>% Matrix(sparse=TRUE)
  pct <- (cts %*% Diagonal(ncol(cts), 1/colSums(cts)))
  lpct <- pct
  lpct@x <- log2(lpct@x)
  2^(-colSums(pct*lpct))
}

filter_cells <- function (sce, idx) { sce[,idx] }

## exclude cells with low diversity
## this often includes both high MT and high Hb(a|b) genes
filter_low_diversity_cells <- function (sce, min_eff_genes=MIN_EFF_GENES) {
  n_eff_genes <- compute_effective_gene_count(sce)
  sce[, n_eff_genes >= min_eff_genes]
}

## attach multimapping genes
attach_multimapping <- function (sce, df) {
  rowData(sce) %<>%
    as_tibble(rownames="rownames") %>%
    left_join(df, by="gene_symbol") %>%
    column_to_rownames(var="rownames") %>%
    DataFrame()
  sce
}

## identify genes that are consistent across datasets
get_consistent_genes <- function (sce_sq, sce_cr) { #}, max_ratio=MAX_RATIO) {
  idx <- intersect(rownames(sce_sq), rownames(sce_cr))
  genes_multimapping <- rowData(sce_sq) %>%
    as_tibble(rownames='rownames') %>%
    filter(has_multimapping) %>%
    pull(rownames)
  setdiff(idx, genes_multimapping)
  # tibble(gene=idx,
  #        cts1=rowSums(counts(sce_sq[idx,])),
  #        cts2=rowSums(counts(sce_cr[idx,]))) %>%
  #   filter(cts1 > 0, cts2 > 0) %>% # exclude undetected genes
  #   filter(cts1/cts2 < max_ratio, cts2/cts1 < max_ratio) %>%
  #   pull(gene)
}
  
filter_features <- function (sce, idx) { sce[idx,] }

## adds PCA to SCE object
add_pca <- function (sce) {
  ## avoid stupid dimnames issue (https://github.com/Bioconductor/SummarizedExperiment/issues/51)
  names(dimnames(counts(sce))) <- NULL
  
  tmp_clusters <- quickCluster(sce)
  sce <- computeSumFactors(sce, clusters=tmp_clusters)
  sce <- logNormCounts(sce)
  dec <- modelGeneVar(sce)
  hvgs <- getTopHVGs(dec, n=N_HVGS)
  runPCA(sce, ncomponents=N_PCS, subset_row=hvgs)
}
```

# Load and Preprocess Data
```{r load_data, message=FALSE, warning=FALSE}
df_counts <- read_csv(SAMPLE_SHEET) %>%
  ## read data files
  mutate(sce_sq=map(file_scutrquant, readRDS),
         sce_10x=map(file_cellranger, read10xCounts),
         tbl_multimap=map(file_multimap, read_multimap_tsv)) %>%
  
  ## adjust cell_ids
  mutate(sce_10x=map2(sample_id, sce_10x, conform_cell_ids)) %>%
  
  ## conform gene_ids
  mutate(sce_sq=map(sce_sq, conform_gene_ids)) %>%
  
  ## filter low-diversity cells
  mutate(sce_sq=map(sce_sq, filter_low_diversity_cells),
         sce_10x=map(sce_10x, filter_low_diversity_cells)) %>%
  
  ## filter to common cells
  mutate(idx_cells=map2(sce_sq, sce_10x, intersect_cell_ids)) %>%
  mutate(sce_sq=map2(sce_sq, idx_cells, filter_cells),
         sce_10x=map2(sce_10x, idx_cells, filter_cells)) %>%
  
  ## attach multimapping genes
  mutate(sce_sq=map2(sce_sq, tbl_multimap, attach_multimapping)) %>%
  
  ## get consistent genes
  mutate(idx_genes=map2(sce_sq, sce_10x, get_consistent_genes))
```

# Analysis
## Plot Gene Counts
```{r plot_gene_counts, fig.width=5, fig.height=4}
plot_umis_per_gene_compare <- function (sce_x, sce_y, idx,
                                        label_x="CellRanger UMI Counts Per Cell",
                                        label_y="scUTRquant UMI Counts Per Cell") {
  idx_common <- intersect(rownames(sce_x), rownames(sce_y))
  df <- tibble(x=rowSums(counts(sce_x[idx_common,])),
               y=rowSums(counts(sce_y[idx_common,])),
               is_consistent=idx_common %in% idx)
  
  ggplot(df, aes(x=x+1, y=y+1, color=is_consistent)) +
    geom_point(alpha=0.3, size=0.1, pch=16) +
    geom_abline(slope=1, intercept=0, linetype='dashed') +
    scale_x_log10() + scale_y_log10() +
    scale_color_manual(values=c('red', 'black')) +
    labs(x=label_x, y=label_y) +
    guides(color=FALSE) +
    theme_minimal_grid()
}

df_counts %>%
  transmute(sample_id, g=pmap(list(sce_x=sce_10x, sce_y=sce_sq, idx=idx_genes), plot_umis_per_gene_compare)) %>%
  deframe() %>% { 
    for (id in names(.)) { 
      ## print locally in this document
      print(.[[id]] + ggtitle(id))
      
      ## save to PDFs
      # ggsave(sprintf("output/fig1B-umis-per-cell-compare-%s.pdf", id), 
      #        .[[id]], width=5, height=4, dpi=300)
    }
  }
```

## Filter Genes and Compute PCA
```{r filter_pca}
df_counts %<>%
  mutate(sce_sq=map2(sce_sq, idx_genes, filter_features),
         sce_10x=map2(sce_10x, idx_genes, filter_features)) %>%

  ## add PCA to SCEs
  mutate(sce_sq=map(sce_sq, add_pca),
         sce_10x=map(sce_10x, add_pca))
```

## Export
```{r export_counts}
saveRDS(df_counts, "data/sce/processed/df_counts.no_multimapping.pca.Rds")
```

# Clustering
## Louvain
```{r compute_louvain_clusters}
cluster_louvain <- function (sce) {
  clusterRows(reducedDim(sce, "PCA"), NNGraphParam(cluster.fun="louvain"))
}

df_counts %>%
  ## cluster
  mutate(clusters_sq=map(sce_sq, cluster_louvain),
         clusters_10x=map(sce_10x, cluster_louvain)) %>%
  
  ## compute adjusted Rand index
  transmute(sample_id, n_cells=map_int(idx_cells, length), 
            ari=map2_dbl(clusters_sq, clusters_10x, aricode::ARI),
            nmi=map2_dbl(clusters_sq, clusters_10x, aricode::NMI),
            nvi=map2_dbl(clusters_sq, clusters_10x, aricode::NVI)) %>%
  
  knitr::kable()
```


---

# Session Info
```{r sesh_info, echo=FALSE}
sessionInfo()
```
