---
title: "Expanded Cell Types - Isoform Changes"
author: "Mervin M Fansler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    code_folding: show
    toc: true
    toc_float: true
---

# Purpose

Here we examine which isoform changes the most when DTU changes are detected.

# Initialization

## Libraries
```{r libs, message=FALSE, warning=FALSE}
library(magrittr)
library(tidyverse)
library(writexl)
library(SingleCellExperiment)
library(Matrix)
library(ggbeeswarm)
```

## Parameters
```{r set_params}
MAX_QVAL=0.05
MIN_L2FC=log2(1.5)
MIN_DWUI=0.10

FILE_SCE="data/sce/merged.txs.full_annot.Rds"
TEST_DTU="data/dtu/20230523-diff-pairwise-wui-tests.min50.boot10k.tsv.gz"
TEST_DGE="data/dge/20230530-diff-pairwise-t-tests.tsv.gz"

celltype_map <- c(
  "Bone Marrow (LSK,LK) - 14/Stem"="HSC",
  "Bone Marrow (LSK,LK) - 12/Ery"="Ery",
  "Bone Marrow (LSK,LK) - 7/Neu"="Neutro",
  "Bone Marrow (LSK,LK) - 6/Baso"="Baso",
  "Bone Marrow (LSK,LK) - 5/Mo"="Mono",
  "Brain - OLG"="ODC",
  "Brain - OPC"="OPC",
  "Brain - ImmN"="imNeur",
  "Brain - mNEUR"="mNeur",
  "Trachea - epithelial cell"="TraEpi",
  "Trachea - mesenchymal cell"="TraMes",
  "Mammary_Gland - luminal epithelial cell of mammary gland"="lumMamm",
  "Mammary_Gland - basal cell"="basMamm",
  "Thymus - immature T cell"="Timm",
  "Spleen - T cell"="TSpleen",
  "Spleen - B cell"="BSpleen",
  "Mammary_Gland - T cell"="TMamm",
  "Mammary_Gland - B cell"="BMamm",
  "Brain - ASC"="Astro",
  "Brain - MG"="MG",
  "Brain - NendC"="NeurEC",
  "Brain - EC"="endo",
  "Embryo - mESC"="mESC",
  "Embryo - MEF"="MEF"
)

test_pairs <- list(
  c("mESC", "MEF"),
  c("HSC", "Ery"),
  c("HSC", "Neutro"),
  c("HSC", "Baso"),
  c("HSC", "Mono"),
  c("Timm", "TSpleen"),
  c("OPC", "ODC"),
  c("imNeur", "mNeur"),
  c("TraMes", "TraEpi"),
  c("lumMamm", "basMamm"),
  c("TSpleen", "TMamm"),
  c("TSpleen", "BSpleen"),
  c("mNeur", "NeurEC"),
  c("mNeur", "Astro"),
  c("mNeur", "ODC"),
  c("mNeur", "MG"),
  c("mNeur", "endo")
)
```

# Data
## Loading
```{r load_data, message=FALSE}
sce <- readRDS(FILE_SCE)
df_dtu <- read_tsv(TEST_DTU) %>%
  dplyr::rename(gene_id=gene)
df_dge <- read_tsv(TEST_DGE)
```

## Preprocessing
### Subset SCE
```{r subset_sce}
colData(sce)['tissue_celltype'] <- colData(sce) %>% 
  { str_c(.$tissue, " - ", .$cell_type) }
idx_cells <- colData(sce) %>% { (.$age != 'old') & (.$tissue_celltype %in% names(celltype_map)) }
idx_txs <- rowData(sce) %>% { (!.$is_blacklisted) & (!.$is_ipa) & (.$atlas.utr_type == "multi") }

sce %<>% `[`(, idx_cells)

colData(sce)['celltype_label'] <- celltype_map[sce$tissue_celltype]

M_cells_celltype <- sce$celltype_label %>% 
  `names<-`(NULL) %>% fac2sparse %>% t

cts_tx_celltype <- counts(sce) %*% M_cells_celltype
df_coldata <- tibble(celltype_label=colnames(M_cells_celltype),
                     n_cells=colSums(M_cells_celltype))

tpm_tx_celltype <- cts_tx_celltype %>%
  { . %*% Diagonal(ncol(.), 1e6/colSums(.)) } %>%
  `colnames<-`(colnames(cts_tx_celltype)) %>%
  as.matrix()


se <- SummarizedExperiment(assays=list(counts=cts_tx_celltype,
                                       tpm=tpm_tx_celltype),
                           colData=df_coldata,
                           rowData=rowData(sce))

se %<>% `[`(idx_txs,)

gene_id2sym <- rowData(se) %>% as_tibble %$% setNames(gene_name, gene_id)
```

```{r compute_sig_tbls}
df_sig_dtu <- df_dtu %>%
  mutate(sig_isoform=qval < MAX_QVAL & abs(stat) >= MIN_DWUI) %>%
  select(gene_id, start, end, sig_isoform)

df_sig_dge <- df_dge %>%
    mutate(sig_gene=qval < MAX_QVAL & abs(log2FC) >= MIN_L2FC) %>%
  select(gene_id, start, end, sig_gene)

df_sig_all <- full_join(df_sig_dtu, df_sig_dge, by=c("gene_id", "start", "end")) %>%
  replace_na(list(sig_isoform=FALSE, sig_gene=FALSE)) %>%
  mutate(sig=ifelse(sig_isoform & sig_gene, "Both", 
                    ifelse(sig_isoform, "DTU", 
                           ifelse(sig_gene, "DGE", "NS"))),
         sig=factor(sig, levels=c("NS", "DTU", "Both", "DGE")),
         test_label=str_c(end, " vs ", start))

df_sig_multi <- left_join(df_sig_dtu, df_sig_dge, by=c("gene_id", "start", "end")) %>%
  replace_na(list(sig_isoform=FALSE, sig_gene=FALSE)) %>%
  mutate(sig=ifelse(sig_isoform & sig_gene, "Both", 
                    ifelse(sig_isoform, "DTU",
                           ifelse(sig_gene, "DGE", "NS"))),
         sig=factor(sig, levels=c("NS", "DTU", "Both", "DGE")),
         test_label=str_c(end, " vs ", start))
```


# Analysis
## Compute TPMs, FCs, and determine site of most change
```{r}
sig_genes <- df_sig_multi %>%
  filter(sig_isoform) %>%
  distinct(gene_id) %$%
  gene_id

sig_gene_test <- df_sig_multi %>%
  filter(sig_isoform) %>%
  distinct(gene_id, start, end)

df_isoforms <- rowData(se) %>% as_tibble %>%
  filter(gene_id %in% sig_genes) %>%
  filter(atlas.pct_utr_no_ipa > 0.1) %>%
  add_count(gene_id, name="n_utrs") %>%
  filter(n_utrs == 2) %>%
  group_by(gene_id) %>%
  mutate(utr_rank=rank(atlas.pct_utr_no_ipa)) %>%
  summarize(tx_weak=transcript_id[utr_rank == 2],
            tx_strong=transcript_id[utr_rank == 1])

df_fcs <- df_sig_multi %>%
  filter(sig_isoform) %>%
  left_join(df_isoforms, by="gene_id") %>%
  filter(!is.na(tx_weak)) %>%
  mutate(tpm_weak_start=map2_dbl(.$tx_weak, .$start, ~ tpm_tx_celltype[.x, .y]),
         tpm_weak_end=map2_dbl(.$tx_weak, .$end, ~ tpm_tx_celltype[.x, .y]),
         tpm_strong_start=map2_dbl(.$tx_strong, .$start, ~ tpm_tx_celltype[.x, .y]),
         tpm_strong_end=map2_dbl(.$tx_strong, .$end, ~ tpm_tx_celltype[.x, .y])) %>%
  mutate(fc_weak=tpm_weak_end/tpm_weak_start,
         fc_strong=tpm_strong_end/tpm_strong_start,
         l2fc_weak=log2(fc_weak),
         l2fc_strong=log2(fc_strong)) %>%
  mutate(site_larger_fc=ifelse(abs(l2fc_strong) >= abs(l2fc_weak), "strong", "weak"),
         celltype_lower_tpm=ifelse(site_larger_fc == "strong",
                                   ifelse(l2fc_strong > 0, "start", "end"),
                                   ifelse(l2fc_weak > 0, "start", "end"))) %>%
  mutate(does_weaker_change_more=case_when(
    celltype_lower_tpm == "start" & site_larger_fc == "strong" ~ tpm_strong_start < tpm_weak_start,
    celltype_lower_tpm == "start" & site_larger_fc == "weak"   ~ tpm_strong_start > tpm_weak_start,
    celltype_lower_tpm == "end" & site_larger_fc == "strong"   ~ tpm_strong_end < tpm_weak_end,
    celltype_lower_tpm == "end" & site_larger_fc == "weak"     ~ tpm_strong_end > tpm_weak_end))
```


## Compare Global "Weak" and "Strong" Sites
The following define weak and strong sites from the atlas, rather than based on the cell type usage.

```{r plots_fc, fig.width=12, fig.height=12}
df_fcs %>%
  ggplot(aes(x=l2fc_strong, y=l2fc_weak)) +
  geom_point() +
  geom_abline(linetype='dashed', color='orchid') +
  facet_wrap(vars(test_label)) +
  theme_bw()

df_fcs %>%
  ggplot(aes(x=abs(l2fc_strong), y=abs(l2fc_weak))) +
  geom_point() +
  geom_abline(linetype='dashed', color='orchid') +
  facet_wrap(vars(test_label)) +
  theme_bw()
```

```{r}
df_fcs %>%
  filter(!is.na(fc_weak), !is.na(fc_strong)) %>%
  group_by(test_label) %>%
  summarize(n_greater=sum(abs(log2(fc_weak)) > abs(log2(fc_strong))), n_total=n(),
            frac_greater=n_greater/n_total)
```

```{r}
df_fcs %>%
  filter(!is.na(fc_weak), !is.na(fc_strong)) %>%
  group_by(test_label, sig) %>%
  summarize(n_greater=sum(fc_weak > fc_strong), n_total=n(),
            frac_greater=n_greater/n_total)
```

## Compare Cell-Type-Specific "Weak" and "Strong" Sites
Here we use the following definition of weak and strong:

For each significant gene-test pair, we determine which site had the most fold-change.
For this site, we determine which cell type it had lower expression. Then we check
whether this site had lower or higher expression than the other site.

```{r}
df_fcs %>%
  group_by(test_label) %>%
  summarize(fraction_weaker_changed_more=mean(does_weaker_change_more))
```


```{r}
df_fcs %>%
  filter(!is.na(does_weaker_change_more)) %>%
  group_by(test_label) %>%
  summarize(fraction_weaker_changed_more=mean(does_weaker_change_more)) %>%
  ggplot(aes(y=test_label, x=fraction_weaker_changed_more)) +
  geom_bar(stat='identity') +
  scale_x_continuous(expand=c(0,0,0,0), limits=c(0,1)) +
  labs(x="Fraction of Weaker Sites with Larger Fold-Change",
       y=NULL) +
  theme_bw()
```

```{r fig.width=2, fig.height=4}
df_fcs %>%
  filter(!is.na(does_weaker_change_more)) %>%
  group_by(test_label) %>%
  summarize(fraction_weaker_changed_more=mean(does_weaker_change_more)) %>%
  ggplot(aes(x="n = 17", y=fraction_weaker_changed_more)) +
  stat_summary(fun='mean', geom='bar', width=0.4, fill='lightgrey', color='black') +
  stat_summary(fun.data=mean_se, geom='errorbar', width=0.3) +
  geom_quasirandom(width=0.1, size=1) +
  scale_y_continuous(limits=c(0,1), expand=expansion(add=c(0,0))) +
  labs(x=NULL, y="Fraction of Weaker Sites with Larger Fold-Change") +
  theme_bw()

ggsave("output/figure4/sup-weaker-isoform-foldchange.pdf", width=2, height=4, dpi=300)
```


---

# Runtime Details
## Session Info
```{r sesh_info, echo=FALSE}
sessionInfo()
```

## Conda Environment
```{bash comment="", echo=FALSE}
if ! command -v conda &> /dev/null
then
  echo "Conda not detected."
elif [ -z "${CONDA_PREFIX}" ]
then
  echo "No active Conda environment."
else
  echo "## Conda Environment YAML"
  conda env export
fi
```
