---
title: "Fansler et al., Nature Communications, Suppl. Data 7"
author: "Sibylle Mitschka"
output: 
  html_document: 
    code_folding: show
    df_print: paged
---

Loading of R packages (not all may be required for this code section). 

```{r}

library(tidyverse)
library(broom)
library(ggpubr)
library(RColorBrewer)
library(writexl)
library(readxl)

```

load data file and generate summarized data table.

```{r message=FALSE, warning=FALSE}


### this data file comprises the gene expression counts and WUI and pct_ipa values for all sgRNA and gene pair conditions

df_WUIess <-readRDS("input_data/df_wui_kd6_essential_ui10.Rds")

### This file provides the cluster identity for a subset of the perturbations (836 genes) which will be used to summarize the data on the cluster level, the clsuter_ids provided in this file do not match the final cluster numbers presented in the paper

target_clus <- read.csv("input_data/df_target_nnclusters_kd6_essential_ui10.csv") %>% 
  mutate(cluster_id_target= as.factor(cluster_id))

### This file provides the names for the clusters and the conversion of the primary cluster id to the final cluster id used in the paper.

names <- read.csv("input_data/names_perturbation_cluster.csv") %>% 
  mutate(cluster_id_target = as.factor(cluster_id_target))

```

```{r}

n_cells_summary<- df_WUIess %>%
  select(target_gene, target_gene_id, sgID_AB, n_cells) %>%
  distinct()

APA_regulator_set <- 
  left_join(target_clus, n_cells_summary, by =c("target_gene", "target_gene_id", "sgID_AB")) %>%
  left_join(names, by= "cluster_id_target") %>%
  dplyr::rename(Gene_symbol =	target_gene,
                Ensembl_ID  = target_gene_id,
                Cluster_id  = order_heatm_new,
                Cluster_name = clus_name) %>%
  dplyr::select(Gene_symbol,Ensembl_ID,sgID_AB, n_cells, Cluster_id, Cluster_name)

write_xlsx(APA_regulator_set, "table_outputs/Suppl_Data7.xlsx")

```

The GO term analyses results presented in sheets 2 and 3 of Suppl. data 7 were added afterwards. Specifically the web interface of the ClusterProfiler tool was used to extarct a list of enriched nd depleted GO terms of the APA regulator set (n= 836 genes) over all tested essential genes in this screen (n= 2,057 genes).


---

# Runtime Details
## Session Info
<details>
```{r sesh_info, echo=FALSE}
sessionInfo()
```
</details>

## Conda Environment
<details>
```{bash comment="", echo=FALSE}
if ! command -v conda &> /dev/null
then
  echo "Conda not detected."
elif [ -z "${CONDA_PREFIX}" ]
then
  echo "No active Conda environment."
else
  echo "## Conda Environment YAML"
  conda env export
fi
```
</details>


