---
title: "DTU vs DGE in 451Lu Data"
author: "Mervin M Fanlser"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    code_folding: show
    toc: true
    toc_float: true
---

# Purpose

Compare the significant differentially expressed genes to the differential LUI
genes in the *Ho et al., 2018* dataset of 451Lu cells with B-Raf inhibitor
resistance.

# Initialization

## Libraries
```{r libs, message=FALSE, warning=FALSE}
library(magrittr)
library(tidyverse)
library(cowplot)
library(org.Hs.eg.db)
```

## Parameters
```{r set_params}
set.seed(20210818)
```

## Functions

```{r methods}

```

# Data
## Loading
```{r load_data, message=FALSE}
df_dge <- read_csv("data/dge/Ho18_Supplemental_Table_S3.csv")

df_dtu <- read_tsv("data/dtu/20211021-451lu-twosample-lui.min50.boot10k.tsv.gz") %>%
  mutate(ensembl_id=str_extract(gene, "^[^.]+"))
```

## Preprocessing
```{r prepare_data}
df_dtu_filtered <- mapIds(org.Hs.eg.db, df_dtu$ensembl_id, "SYMBOL", "ENSEMBL") %>%
  enframe(name="ensembl_id", value="gene_symbol") %>%
  full_join(df_dtu, by="ensembl_id") %>%
  filter(qval < 0.05, abs(stat) > 0.15)

df_dge_filtered <- df_dge %>% 
  filter(padj < 0.05, abs(log2FoldChange) >= 1)
```

# Analysis

## Common Genes

```{r intersect_genes, fig.width=5, fig.height=4}
intersect(df_dtu_filtered$gene_symbol, df_dge_filtered$Gene)
```


# Conclusion

It appears there is almost no overlap, however, it is unclear if this gene list 
from the SAKE paper is pre-filtered in some way. 

---

# Runtime Details
## Session Info
```{r sesh_info, echo=FALSE}
sessionInfo()
```

## Conda Environment
```{bash comment="", echo=FALSE}
if ! command -v conda &> /dev/null
then
  echo "Conda not detected."
elif [ -z "${CONDA_PREFIX}" ]
then
  echo "No active Conda environment."
else
  echo "## Conda Environment YAML"
  conda env export
fi
```
