---
title: "t-Test Results - Erythrocyte Trajectory"
author: "Mervin M Fansler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    code_folding: show
    toc: true
    toc_float: true
    df_print: paged
---

# Purpose

This analysis processes the two-sample DGE test results to identify significant
gene expression changes in the HSC to Erythrocyte trajectory from the *Dahlin et al., 2018*
dataset. Significant genes are then plotted as a heat map.

# Initialization

## Libraries
```{r libs, message=FALSE, warning=FALSE}
library(magrittr)
library(tidyverse)
library(cowplot)
library(SingleCellExperiment)
library(Matrix)
library(matrixStats)
library(pheatmap)
```

## Parameters
```{r set_params}
set.seed(20211020)
MAX_QVAL=0.05
MIN_FC=1.5
MIN_L2FC=log2(MIN_FC)

TSV_TESTS="data/dge/20230530-ery-trajectory-pairwise-t-tests.tsv.gz"
SCE_GENES="data/sce/hspcs.genes.full_annot.Rds"

PDF_OUT_UTRTYPE=sprintf("output/figure4/fig4d-dge-fc%0.1f-hsc-ery-path-utr-type.pdf", MIN_FC)
PDF_OUT_DIRECTION=sprintf("output/figure4/fig4d-dge-fc%0.1f-hsc-ery-path-direction.pdf", MIN_FC)
PDF_OUT_BOTH=sprintf("output/figure4/fig4d-dge-fc%0.1f-hsc-ery-path-utr-type-direction.pdf", MIN_FC)
```

## Functions

```{r methods}

```

# Data
## Loading
```{r load_data, message=FALSE}
df_tests <- read_tsv(TSV_TESTS, col_types='ccccddd') %>%
  mutate(test=str_c(end, " vs ", start),
         qval=p.adjust(pval, method='BH'),
         sig=(qval < MAX_QVAL) & (abs(log2FC) >= MIN_L2FC))

sce <- readRDS(SCE_GENES)
```

## Preprocessing
### Compute Summary Tables
```{r compute_summaries}
gid2utr_type <- rowData(sce) %>% as_tibble %>% select(gene_id, atlas.utr_type) %>% deframe

df_summary <- df_tests %>%
  mutate(utr_type=gid2utr_type[gene_id]) %>%
  group_by(test) %>%
  summarize(n_tested=n(),
            n_decrease=sum(sig & log2FC < 0, na.rm=TRUE),
            n_increase=sum(sig & log2FC > 0, na.rm=TRUE),
            pct_decrease=100*n_decrease/n_tested,
            pct_increase=100*n_increase/n_tested,
            n_multi=sum(sig & utr_type == 'multi'),
            n_single=sum(sig & utr_type == 'single'),
            .groups='drop')

df_sig_genes <- df_tests %>%
  #filter(sig) %>%
  group_by(gene_symbol, gene_id) %>%
  summarize(max_test=test[which.max(abs(log2FC))],
            max_log2FC=log2FC[which.max(abs(log2FC))],
            mean_log2FC=mean(log2FC),
            # end_log2FC=log2FC[test == "12/Ery vs 14/Stem"],
            is_sig=any(sig),
            .groups='drop') %>%
  filter(is_sig) %>%
  mutate(utr_type=gid2utr_type[gene_id])

genes_increasing <- filter(df_sig_genes, mean_log2FC > 0)$gene_id
genes_decreasing <- filter(df_sig_genes, mean_log2FC < 0)$gene_id

genes_multi  <- filter(df_sig_genes, utr_type == "multi")$gene_id
genes_single <- filter(df_sig_genes, utr_type == "single")$gene_id

genes_increasing_multi <- filter(df_sig_genes, mean_log2FC > 0, utr_type == "multi")$gene_id
genes_decreasing_multi <- filter(df_sig_genes, mean_log2FC < 0, utr_type == "multi")$gene_id

genes_increasing_single <- filter(df_sig_genes, mean_log2FC > 0, utr_type == "single")$gene_id
genes_decreasing_single <- filter(df_sig_genes, mean_log2FC < 0, utr_type == "single")$gene_id
```

### Compute Significant DGEs
```{r compute_luis}
celltypes <- unique(c(df_tests$start, df_tests$end))

## filter cells
sce %<>% `[`(, .$clusters %in% celltypes)

## design matrices
M_cell_celltype <- t(fac2sparse(sce$clusters))[,celltypes]

## compute tpms
tpm_gene_celltype <- (assay(sce, "normcounts") %*% M_cell_celltype) %>%
  ## normalize to tpm
  { . %*% Diagonal(ncol(.), 1e6/colSums(.), names=TRUE) } %>%
  as.matrix

## filter sig genes
idx_sig_genes <- rownames(tpm_gene_celltype) %in% df_sig_genes$gene_id
tpm_gene_celltype %<>% `[`(idx_sig_genes,)

ztpm_gene_celltype <- t(tpm_gene_celltype) %>% scale %>% t
```



### Compute Orderings
```{r compute_gene_orders}
wt_ztpm <- ztpm_gene_celltype %>% 
  { setNames(as.numeric(. %*% seq(-1,1, length.out=ncol(.))), rownames(.)) }

## split by direction
idx_increasing <- which(names(wt_ztpm) %in% genes_increasing) %>% `[`(order(wt_ztpm[.]))
idx_decreasing <- which(names(wt_ztpm) %in% genes_decreasing) %>% `[`(order(wt_ztpm[.]))

idx_direction <- c(idx_increasing, idx_decreasing)

## split by UTR type
idx_multi  <- which(names(wt_ztpm) %in% genes_multi) %>% `[`(order(wt_ztpm[.]))
idx_single <- which(names(wt_ztpm) %in% genes_single) %>% `[`(order(wt_ztpm[.]))

idx_utr_type <- c(idx_multi, idx_single)
df_rowdata_utr_type <- data.frame("Type"=rep(c("Multi-UTR", "Single-UTR"), 
                                                 c(length(idx_multi), length(idx_single))),
                                  row.names=names(wt_ztpm)[idx_utr_type])
  
## split by direction and UTR type
idx_increasing_multi  <- which(names(wt_ztpm) %in% genes_increasing_multi) %>% `[`(order(wt_ztpm[.]))
idx_decreasing_multi  <- which(names(wt_ztpm) %in% genes_decreasing_multi) %>% `[`(order(wt_ztpm[.]))
idx_increasing_single <- which(names(wt_ztpm) %in% genes_increasing_single) %>% `[`(order(wt_ztpm[.]))
idx_decreasing_single <- which(names(wt_ztpm) %in% genes_decreasing_single) %>% `[`(order(wt_ztpm[.]))

idx_direction_utr_type <- c(idx_increasing_multi, idx_decreasing_multi,
                            idx_increasing_single, idx_decreasing_single)
df_rowdata_direction_utr_type <- data.frame("Type"=rep(c("Multi-UTR", "Single-UTR"), 
                                                       c(length(idx_multi), length(idx_single))),
                                            "Direction"=rep(c("Increasing", "Decreasing",
                                                              "Increasing", "Decreasing"),
                                                            c(length(idx_increasing_multi), 
                                                              length(idx_decreasing_multi), 
                                                              length(idx_increasing_single), 
                                                              length(idx_decreasing_single))),
                                            row.names=names(wt_ztpm)[idx_direction_utr_type])
```

# Summary Overview
We have `r length(idx_increasing_multi)` multi-UTR genes increasing and `r length(idx_decreasing_multi)` decreasing. We have `r length(idx_increasing_single)` single-UTR genes increasing and `r length(idx_decreasing_single)` decreasing.


## Significant Tests

```{r sig_tests}
df_summary
```


## Significant Genes
```{r sig_genes}
df_sig_genes %>%
  summarize(n_increasing=sum(mean_log2FC > 0),
            n_decreasing=sum(mean_log2FC < 0)) %>%
  knitr::kable()

df_sig_genes %>%
  group_by(max_test) %>%
  summarize(n_increasing=sum(mean_log2FC > 0),
            n_decreasing=sum(mean_log2FC < 0),
            .groups='drop') %>%
  knitr::kable()
```


# Plot Heatmap
## UTR Type
### Z-score
```{r fig_heatmap_ztpm_type, fig.width=4, fig.height=4}
NCOLORS=100
breaks.point <- seq(-3, 3, length.out=NCOLORS + 1)

pheatmap(ztpm_gene_celltype[idx_utr_type,], 
         cluster_rows=FALSE, cluster_cols=FALSE, 
         labels_col=c("HSC", "Prog1", "Prog2", "MEP", "ProE", "Baso1", "Baso2", "Ery"),
         show_rownames=FALSE, scale='none', fontsize=8, 
         annotation_row=df_rowdata_utr_type,
         color=colorRampPalette(c("#336BA8", "white", "#D62125"))(NCOLORS),
         breaks=breaks.point,
         angle_col=0,
         filename=PDF_OUT_UTRTYPE,
         width=8, base_height=6)
```

## Direction
### Z-score
```{r fig_heatmap_ztpm_dir, fig.width=4, fig.height=4}
NCOLORS=100
breaks.point <- seq(-3, 3, length.out=NCOLORS + 1)

pheatmap(ztpm_gene_celltype[rev(idx_direction),], 
         cluster_rows=FALSE, cluster_cols=FALSE, 
         labels_col=c("HSC", "Prog1", "Prog2", "MEP", "ProE", "Baso1", "Baso2", "Ery"),
         show_rownames=FALSE, scale='none', fontsize=8, 
         color=colorRampPalette(c("#336BA8", "white", "#D62125"))(NCOLORS),
         breaks=breaks.point,
         angle_col=0,
         filename=PDF_OUT_DIRECTION,
         width=8, base_height=6)
```

## Direction & UTR Type
### Z-score
```{r fig_heatmap_ztpm_dir_type, fig.width=4, fig.height=4}
NCOLORS=100
breaks.point <- seq(-3, 3, length.out=NCOLORS + 1)

pheatmap(ztpm_gene_celltype[idx_direction_utr_type,], 
         cluster_rows=FALSE, cluster_cols=FALSE, 
         labels_col=c("HSC", "Prog1", "Prog2", "MEP", "ProE", "Baso1", "Baso2", "Ery"),
         show_rownames=FALSE, scale='none', fontsize=8, 
         annotation_row=df_rowdata_direction_utr_type,
         color=colorRampPalette(c("#336BA8", "white", "#D62125"))(NCOLORS),
         breaks=breaks.point,
         angle_col=0, 
         filename=PDF_OUT_BOTH,
         width=8, base_height=6)
```

### TPM
```{r fig_heatmap_tpm, fig.width=4, fig.height=4}
NCOLORS=100
breaks.point <- seq(0, 5, length.out=NCOLORS + 1)

pheatmap(log10(tpm_gene_celltype[rev(idx_direction),]), 
         cluster_rows=FALSE, cluster_cols=FALSE, 
         labels_col=c("HSC", "Prog1", "Prog2", "MEP", "ProE", "Baso1", "Baso2", "Ery"),
         show_rownames=FALSE, scale='none', fontsize=8,
         color=colorRampPalette(viridis::viridis(3))(NCOLORS),
         breaks=breaks.point, 
         angle_col=0)
```



---

# Runtime Details
## Session Info
```{r sesh_info, echo=FALSE}
sessionInfo()
```

## Conda Environment
```{bash comment="", echo=FALSE}
if ! command -v conda &> /dev/null
then
  echo "Conda not detected."
elif [ -z "${CONDA_PREFIX}" ]
then
  echo "No active Conda environment."
else
  echo "## Conda Environment YAML"
  conda env export
fi
```
